---
title: "Workflow-v2"
author: "Anna Willoughby"
date: "2025-05-22"
output: pdf_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library('deSolve') # solving ordinary differential equations
library(dplyr) # data wrangling
library(ggplot2) # pretty plots
library(lhs) # latin hypercube sampling
set.seed(1993)
```

## Model

In this document we seek to use a birth-death-immigration process to model the number of ectoparasites on a host.A full description of the model is at https://github.com/DrakeLab/willoughby-grooming-model.

We start by writing functions for the birth rate, death rate, and immigration rate.
In all cases, we let the state variable ($x$) be the first parameter.
We also set some default parameters to make things simpler later.


```{r, functions}
birthrate <- function(x, b0, theta) b0*x*(x/(theta+x)) # function for birth rate with Allee effect (theta)

# we are going to explore different mortality rates through grooming scenarios: 
## PROGRAMMED 
## G is some threshold of ectos that the host can groom per time period
deathrate_prg <- function(x, G) {
  pmin(1, G / x)  # cap at 1 to avoid unrealistic death rates > 1
} 

## STIMULUS - Driven 
## dmax =  maximum grooming-induced per capita death rate
## 𝑘 = half-saturation constant — the population size at which half of dmax is reached
# deathrate_stim <- function(x, d0, dmax, k) (d0 + (dmax * x) / (k + x))*x

immigrationrate <- function (x, iota) iota # function for immigration rate (iota)

# set initial conditions parameters

## birth rate is standard across all scenarios, right now as 100 flea births per year (0.27 per day)
## we are going to vary parameters for:  
theta_vals <- seq(0, 10, by = 1)
iota_vals <- seq(0, 10, by = 1) # 
# variation in the max threshold 
G_vals <- seq(0,100, by = 1)

n_samples <- 500  # number of parameter combos
lhs_matrix <- randomLHS(n_samples, 3)  # for theta, G, iota

# Rescale each column to your parameter ranges
theta_vals <- lhs_matrix[,1] * 10        #  allee effect (none to high (n = 10))
G_vals     <- lhs_matrix[,2] * 100       # 0 to 100
iota_vals  <- lhs_matrix[,3] * 10        # immigration (none to high (n = 10))

# Construct parameter set
param_grid <- data.frame(
  b0 = 0.27, # constant
  theta = theta_vals,
  G = G_vals,
  iota = iota_vals,
  initial_pop = 2 # constant
)
```

## Stochastic Model

Now, we want to simulate a birth-death-immigration process.
We start with some hyperparameters of the simulation.

```{r set some simulation rules}
num_simulations <- 1  # Number of times to run the simulation
# max.events <- 1000 # maximum events
max.time <- 730 # 2 year max for a squirrel  
```

Here we simulate the process

```{r run stochastic models}
system.time({
  for (i in 1:10) {

simulation_pg <- vector("list", nrow(param_grid))  # one sim per parameter set

for (i in 1:nrow(param_grid)) {
  
  # Extract parameter values for this run
  parms <- param_grid[i, ]
  
  # Initialize
  events <- 0
  time <- 0
  x <- parms$initial_pop
  
  # Store simulation output
  simulation_results <- data.frame(
    event = numeric(),
    time = numeric(),
    population = numeric(), 
    event_type = character(),
    sim_id = numeric(),
    param_id = numeric() 
  )

  while (time[length(time)] <= max.time) {
    events <- events + 1
    current.size <- x[length(x)]
    current.time <- time[length(time)]
    
    # Calculate rates using current parasite load and this parameter set
    b <- birthrate(current.size, b0 = parms$b0, theta = parms$theta)
    d <- deathrate_prg(current.size, G = parms$G)
    iota <- immigrationrate(current.size, iota = parms$iota)
    
    total.rate <- b + d + iota
    if (total.rate == 0 || is.na(total.rate)) break

    increment.time <- -log(1 - runif(1)) / total.rate
    
    # Choose which event occurs
    event_probs <- c(d, b, iota) / total.rate
    event_outcome <- sample(c("death", "birth", "immigration"), size = 1, prob = event_probs)
    change <- ifelse(event_outcome == "death", -1,
              ifelse(event_outcome == "birth", 1, 1))  # immigration = +1
    
    new_size <- max(0, current.size + change)
    new_time <- current.time + increment.time

    # Append row
    simulation_results <- rbind(simulation_results, data.frame(
      event = events,
      time = new_time,
      population = new_size,
      event_type = event_outcome,
      sim_id = 1,
      param_id = 1
    ))

    time <- c(time, new_time)
    x <- c(x, new_size)
  }

  # Store result for this parameter set
  simulation_pg[[i]] <- simulation_results
}

 }
})
saveRDS(simulation_pg, "data/simulation_pg.Rdata")


```

Now, I need a way to sample from these data.

```{r plot stochastic model output}


# make a plot with all the simulations 
ggplot(results0, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(x = "Time", y = "Population Size", color = "Simulation ID", 
       caption = ) +
  theme_classic() +
  theme(legend.position = "none")  # hide legend cause too many lines
ggplot(results1, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(caption = "Figure X. Ectoparasite population size across time (100 years)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none", plot.caption = element_text(hjust=0.5, size=rel(1.2)))  # hide legend cause too many lines
ggplot(results2, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(caption= "Figure X. Ectoparasite population size across time (100 years)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none", plot.caption = element_text(hjust=0.5, size=rel(1.2)))  # hide legend cause too many lines
ggplot(results3, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(x = "Time", y = "Population Size", color = "Simulation ID", 
       caption="Figure X. Ectoparasite population size across time (100 years)") + 
  theme_classic() +
  theme(legend.position = "none", plot.caption = element_text(hjust=0.5, size=rel(1.2)))  # hide legend cause too many lines

```

```{r calculate traits - total pop}
# FLEA POPULATION TRAITS - calculate relevant traits 

## values for one simulation
### Total Flea Population: the count of fleas during the entire simulation run 

#### Function to calculate total flea population from immigration & birth events
calculate_flea_population <- function(simulation_list) {
  sapply(simulation_list, function(sim_df) {
    sum(sim_df$event_type == "immigration") + sum(sim_df$event_type == "birth")
  })
}

#### Apply function to all models
flea_total_pop_lists <- lapply(all_simulations, calculate_flea_population)

#### Plot histograms for each model as facet
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid
for (i in seq_along(flea_total_pop_lists)) {
  hist(flea_total_pop_lists[[i]], 
       main = paste("Total Flea Pop -", names(flea_total_pop_lists)[i]), 
       xlab = "Total Fleas", 
       col = "lightblue", 
       border = "black")
}
par(mfrow = c(1,1))  # Reset plot layout

#### Convert list to long-format dataframe for ggplot
flea_data <- bind_rows(
  lapply(names(flea_total_pop_lists), function(model) {
    data.frame(Total_Fleas = flea_total_pop_lists[[model]], Model = model)
  })
)

#### Plot histogram using ggplot2 for each model as overlapping on same grid
ggplot(flea_data, aes(x = Total_Fleas, fill = Model)) +
  geom_histogram(aes(y = ..density..), position = "identity", alpha = 0.4, bins = 30) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9","grey", "darkmagenta")) +  # Custom colors
  labs(title = "Total Flea Population Across Simulations",
       x = "Total Fleas from Immigration & Birth",
       y = "Density") +
  theme_minimal()

```

```{r calculate traits - snapshot pops}
### Snapshot/Sampling Event Flea Population: the count of fleas at one snapshot (e.g., end of day)

#### create function round time to the nearest integer, keep only the last observation per time unit
end_of_day_flea_pop <-function(sim){
  sim %>%  # Apply operations to the dataset
  mutate(time_unit = floor(time)) %>%  # Convert time to integer units
  group_by(time_unit) %>%
  slice_tail(n = 1) %>%  # Get the last observation per unit
  ungroup() %>%
  select(time_unit, population)  # Keep relevant columns
}

# apply the end of day calculation across simulations 
sim0_end_of_day_pops <- lapply(simulation0, end_of_day_flea_pop)
sim1_end_of_day_pops <- lapply(simulation1, end_of_day_flea_pop)
sim2_end_of_day_pops <- lapply(simulation2, end_of_day_flea_pop)
sim3_end_of_day_pops <- lapply(simulation3, end_of_day_flea_pop)

# create a function to sample only a portion of the days (e.g. last 25%)
subset_simulation <- function(sim_df, range_percent) {
  # Ensure range_percent is a valid numeric vector of length 2 (e.g., c(0, 10) or c(75, 100))
  if (length(range_percent) != 2 || any(range_percent < 0) || any(range_percent > 100)) {
    stop("range_percent must be a numeric vector of length 2 between 0 and 100.")
  }
  
  # Get the start and end time from the simulation
  min_time <- min(sim_df$time_unit)
  max_time <- max(sim_df$time_unit)

  # Compute the time range corresponding to the percentage range
  time_start <- min_time + (range_percent[1] / 100) * (max_time - min_time)
  time_end <- min_time + (range_percent[2] / 100) * (max_time - min_time)

  # Subset the dataframe based on the calculated time range
  sim_df %>%
    filter(time_unit >= time_start & time_unit <= time_end)
}

# limit population snapshots to latter 75% (eliminating burn-in period)
sim0_eod_pop_last75 <- lapply(sim0_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim0_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim0_eod_pop_last75), function(i) {
    sim0_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim0_eod_pop_last75_combined$model <- "0 - basic"
sim0_eod_pop_last75_combined$sim_id <- paste("0-", sim0_eod_pop_last75_combined$sim_id, sep = "")
sim1_eod_pop_last75 <- lapply(sim1_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim1_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim1_eod_pop_last75), function(i) {
    sim1_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim1_eod_pop_last75_combined$model <- "1 - allee"
sim1_eod_pop_last75_combined$sim_id <- paste("1-", sim1_eod_pop_last75_combined$sim_id, sep = "")

sim2_eod_pop_last75 <- lapply(sim2_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim2_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim2_eod_pop_last75), function(i) {
    sim2_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim2_eod_pop_last75_combined$model = "2 - resources"
sim2_eod_pop_last75_combined$sim_id <- paste("2-", sim2_eod_pop_last75_combined$sim_id, sep = "")

sim3_eod_pop_last75 <- lapply(sim3_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim3_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim3_eod_pop_last75), function(i) {
    sim3_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim3_eod_pop_last75_combined$model = "3 - complex"
sim3_eod_pop_last75_combined$sim_id <- paste("3-", sim3_eod_pop_last75_combined$sim_id, sep = "")

eod_pop_last75 <- rbind(sim0_eod_pop_last75_combined, 
                        sim1_eod_pop_last75_combined, 
                        sim2_eod_pop_last75_combined, 
                        sim3_eod_pop_last75_combined)
# plot the histogram 
# hist(sim0_eod_pop_last75[[1]]$population)
ggplot(eod_pop_last75, aes(x = population, group = sim_id, color = model)) +
 # geom_freqpoly(binwidth = 10, size = 1, alpha = 0.5) +  # stepbins like a histogram
  geom_histogram(alpha = 0.2, aes(y = ..ncount..), position = 'identity', bins = 10) + 
 # geom_bar(pos="dodge") +
  scale_x_continuous(limits = c(0, 50)) +
 # scale_y_continuous(limits = c(0, 0.15)) +
  labs(title = "Distribution of Population Sizes for last 75% of simulation (n = 541 time points) ",
       x = "Population Size",
       y = "Frequency") +
  theme_classic() + 
 facet_grid(model ~ .)
```

```{r calculate aggregation metrics}

```
