---
title: "Workflow-v2"
author: "Anna Willoughby"
date: "2025-05-22"
output: pdf_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library('deSolve') # solving ordinary differential equations
library(dplyr) # data wrangling
library(ggplot2) # pretty plots
library(lhs) # latin hypercube sampling
library(tidyr)
set.seed(1993)
```
```{r set up parallellization}
library(parallel) # for running in parallel 
# define cluster 
# create cluster object
cl <- makeCluster(10) # 10 cores on home desktop
```



## Model

In this document we seek to use a birth-death-immigration process to model the number of ectoparasites on a host. A full description of the model is at https://github.com/DrakeLab/willoughby-grooming-model.

We start by writing functions for the birth rate, death rate, and immigration rate.
In all cases, we let the state variable ($x$) be the first parameter.
We also set some default parameters to make things simpler later.


```{r, functions}

# function for birth rate with Allee effect (theta)
birthrate <- function(x, b0, theta, K){
  b0*x*(x/(theta+x)) 
  b * pmax(0, 1 - x / K)
}

immigrationrate <- function (x, iota) iota # function for immigration rate (iota)

# read in all the death rates 

## Define all Grooming Scenarios: 
### PROGRAMMED 
groom_prg_scenarios <- data.frame(scenario = seq(1, 100, length.out = 100), # 100 scenarios 
                                  G_vals = seq(1, 100, length.out = 100)) # grooming efficiency
#### Generate a function for grooming deathrate based on pop size 
groomrate_prg <- function(x, scenario){
  groom_prg_scenario <- groom_prg_scenarios %>% dplyr::filter(scenario == scenario) # filter to scenario 
  fleas_groomed <- pmin(x, groom_prg_scenario[1,2]) # how much grooming will happen
  death_rate <- ifelse(x == 0, 0, # avoids NaNs
                                  fleas_groomed / x)
  death_rate
}

deathrate_dd <- function(x, dmax = 1, x_mid = 100, s = 20) {
     dmax / (1 + exp(-(x - x_mid) / s))
}
# set initial conditions parameters

## birth rate is standard across all scenarios, right now as 100 flea births per year (0.27 per day)
## we are going to vary parameters for:  
theta_vals <- seq(0, 10, by = 1)
iota_vals <- seq(0, 10, by = 1) 
scenario_vals <- seq(1,100, by = 1)

n_samples <- 500  # number of parameter combos
lhs_matrix <- randomLHS(n_samples, 3)  # for theta, iota

# Rescale each column to your parameter ranges
theta_vals <- lhs_matrix[,1] * 10        #  allee effect (none to high (n = 10))
iota_vals  <- lhs_matrix[,2] * 10        # immigration (none to high (n = 10))
scenario_vals <- floor(lhs_matrix[,3] * 100)    # which grooming scenario 

# Construct parameter set
param_grid <- data.frame(
  b0 = 0.27, # constant
  theta = theta_vals,
  iota = iota_vals,
  scenario = scenario_vals,
  initial_pop = 2 # constant
)

```

## Stochastic Model

Now, we want to simulate a birth-death-immigration process.
We start with some hyperparameters of the simulation.

```{r set some simulation rules}
num_simulations <- 1  # Number of times to run the simulation
# max.events <- 1000 # maximum events
max.time <- 730 # 2 year max for a squirrel  
```

Here we simulate the process

```{r run stochastic models}
# g_scenario <- read.csv(file = "data/grooming_deathrates.csv")
# groom_prog <- g_scenario %>% dplyr::filter(groom_type == "programmed")

# Define running the whole model as a function 
system.time({

  cl <- makeCluster(10)

  clusterExport(cl, c(
    "param_grid", "max.time",
    "birthrate", "groomrate_prg", "deathrate_dd", "immigrationrate"
  ))

  task_grid <- expand.grid(
    param_id = 1:nrow(param_grid),
    sim_id   = 1:10
  )

  simulation_pg <- parLapply(cl, 1:nrow(task_grid), function(task_i) {

    param_id <- task_grid$param_id[task_i]
    sim_id   <- task_grid$sim_id[task_i]
    parms <- param_grid[param_id, ]

    events <- 0
    time <- 0
    x <- parms$initial_pop

    out_list <- list()

    while (time[length(time)] <= max.time) {

      events <- events + 1
      current.size <- x[length(x)]
      current.time <- time[length(time)]

      b <- birthrate(current.size, b0 = parms$b0, theta = parms$theta, K = 300)
      d <- groomrate_prg(current.size, parms$scenario) +
           deathrate_dd(current.size)
      iota <- immigrationrate(current.size, iota = parms$iota)

      total.rate <- b + d + iota

      if (is.na(total.rate) || total.rate <= 0) break

      increment.time <- -log(1 - runif(1)) / total.rate

      event_probs <- c(d, b, iota) / total.rate
      event_outcome <- sample(
        c("death", "birth", "immigration"),
        size = 1,
        prob = event_probs
      )

      change <- ifelse(event_outcome == "death", -1, 1)

      new_size <- max(0, current.size + change)
      new_time <- current.time + increment.time

      out_list[[events]] <- data.frame(
        event = events,
        time = new_time,
        population = new_size,
        event_type = event_outcome,
        sim_id = sim_id,
        param_id = param_id
      )

      time <- c(time, new_time)
      x <- c(x, new_size)
    }

    if (length(out_list) == 0) {
      data.frame(
        event = numeric(),
        time = numeric(),
        population = numeric(),
        event_type = character(),
        sim_id = numeric(),
        param_id = numeric()
      )
    } else {
      do.call(rbind, out_list)
    }
  })

  stopCluster(cl)
})

saveRDS(simulation_pg, "data/simulation_pg.Rdata")


```
Now, I need a way to sample from these data.

```{r plot stochastic model output}
# make a plot with all the simulations 
ggplot(results0, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(x = "Time", y = "Population Size", color = "Simulation ID", 
       caption = ) +
  theme_classic() +
  theme(legend.position = "none")  # hide legend cause too many lines
ggplot(results1, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(caption = "Figure X. Ectoparasite population size across time (100 years)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none", plot.caption = element_text(hjust=0.5, size=rel(1.2)))  # hide legend cause too many lines
ggplot(results2, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(caption= "Figure X. Ectoparasite population size across time (100 years)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none", plot.caption = element_text(hjust=0.5, size=rel(1.2)))  # hide legend cause too many lines
ggplot(results3, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(x = "Time", y = "Population Size", color = "Simulation ID", 
       caption="Figure X. Ectoparasite population size across time (100 years)") + 
  theme_classic() +
  theme(legend.position = "none", plot.caption = element_text(hjust=0.5, size=rel(1.2)))  # hide legend cause too many lines
```

```{r calculate traits - total pop}
# FLEA POPULATION TRAITS - calculate relevant traits 

## values for one simulation
### Total Flea Population: the count of fleas during the entire simulation run 

#### Function to calculate total flea population from immigration & birth events
calculate_flea_population <- function(simulation_list) {
  sapply(simulation_list, function(sim_df) {
    sum(sim_df$event_type == "immigration") + sum(sim_df$event_type == "birth")
  })
}

#### Apply function to all models
flea_total_pop_lists <- lapply(all_simulations, calculate_flea_population)

#### Plot histograms for each model as facet
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid
for (i in seq_along(flea_total_pop_lists)) {
  hist(flea_total_pop_lists[[i]], 
       main = paste("Total Flea Pop -", names(flea_total_pop_lists)[i]), 
       xlab = "Total Fleas", 
       col = "lightblue", 
       border = "black")
}
par(mfrow = c(1,1))  # Reset plot layout

#### Convert list to long-format dataframe for ggplot
flea_data <- bind_rows(
  lapply(names(flea_total_pop_lists), function(model) {
    data.frame(Total_Fleas = flea_total_pop_lists[[model]], Model = model)
  })
)

#### Plot histogram using ggplot2 for each model as overlapping on same grid
ggplot(flea_data, aes(x = Total_Fleas, fill = Model)) +
  geom_histogram(aes(y = ..density..), position = "identity", alpha = 0.4, bins = 30) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9","grey", "darkmagenta")) +  # Custom colors
  labs(title = "Total Flea Population Across Simulations",
       x = "Total Fleas from Immigration & Birth",
       y = "Density") +
  theme_minimal()

```

```{r calculate traits - snapshot pops}
### Snapshot/Sampling Event Flea Population: the count of fleas at one snapshot (e.g., end of day)

#### create function round time to the nearest integer, keep only the last observation per time unit
end_of_day_flea_pop <-function(sim){
  sim %>%  # Apply operations to the dataset
  mutate(time_unit = floor(time)) %>%  # Convert time to integer units
  group_by(time_unit) %>%
  slice_tail(n = 1) %>%  # Get the last observation per unit
  ungroup() %>%
  select(time_unit, population)  # Keep relevant columns
}

# apply the end of day calculation across simulations 
sim0_end_of_day_pops <- lapply(simulation0, end_of_day_flea_pop)
sim1_end_of_day_pops <- lapply(simulation1, end_of_day_flea_pop)
sim2_end_of_day_pops <- lapply(simulation2, end_of_day_flea_pop)
sim3_end_of_day_pops <- lapply(simulation3, end_of_day_flea_pop)

# create a function to sample only a portion of the days (e.g. last 25%)
subset_simulation <- function(sim_df, range_percent) {
  # Ensure range_percent is a valid numeric vector of length 2 (e.g., c(0, 10) or c(75, 100))
  if (length(range_percent) != 2 || any(range_percent < 0) || any(range_percent > 100)) {
    stop("range_percent must be a numeric vector of length 2 between 0 and 100.")
  }
  
  # Get the start and end time from the simulation
  min_time <- min(sim_df$time_unit)
  max_time <- max(sim_df$time_unit)

  # Compute the time range corresponding to the percentage range
  time_start <- min_time + (range_percent[1] / 100) * (max_time - min_time)
  time_end <- min_time + (range_percent[2] / 100) * (max_time - min_time)

  # Subset the dataframe based on the calculated time range
  sim_df %>%
    filter(time_unit >= time_start & time_unit <= time_end)
}

# limit population snapshots to latter 75% (eliminating burn-in period)
sim0_eod_pop_last75 <- lapply(sim0_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim0_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim0_eod_pop_last75), function(i) {
    sim0_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim0_eod_pop_last75_combined$model <- "0 - basic"
sim0_eod_pop_last75_combined$sim_id <- paste("0-", sim0_eod_pop_last75_combined$sim_id, sep = "")
sim1_eod_pop_last75 <- lapply(sim1_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim1_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim1_eod_pop_last75), function(i) {
    sim1_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim1_eod_pop_last75_combined$model <- "1 - allee"
sim1_eod_pop_last75_combined$sim_id <- paste("1-", sim1_eod_pop_last75_combined$sim_id, sep = "")

sim2_eod_pop_last75 <- lapply(sim2_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim2_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim2_eod_pop_last75), function(i) {
    sim2_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim2_eod_pop_last75_combined$model = "2 - resources"
sim2_eod_pop_last75_combined$sim_id <- paste("2-", sim2_eod_pop_last75_combined$sim_id, sep = "")

sim3_eod_pop_last75 <- lapply(sim3_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim3_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim3_eod_pop_last75), function(i) {
    sim3_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim3_eod_pop_last75_combined$model = "3 - complex"
sim3_eod_pop_last75_combined$sim_id <- paste("3-", sim3_eod_pop_last75_combined$sim_id, sep = "")

eod_pop_last75 <- rbind(sim0_eod_pop_last75_combined, 
                        sim1_eod_pop_last75_combined, 
                        sim2_eod_pop_last75_combined, 
                        sim3_eod_pop_last75_combined)
# plot the histogram 
# hist(sim0_eod_pop_last75[[1]]$population)
ggplot(eod_pop_last75, aes(x = population, group = sim_id, color = model)) +
 # geom_freqpoly(binwidth = 10, size = 1, alpha = 0.5) +  # stepbins like a histogram
  geom_histogram(alpha = 0.2, aes(y = ..ncount..), position = 'identity', bins = 10) + 
 # geom_bar(pos="dodge") +
  scale_x_continuous(limits = c(0, 50)) +
 # scale_y_continuous(limits = c(0, 0.15)) +
  labs(title = "Distribution of Population Sizes for last 75% of simulation (n = 541 time points) ",
       x = "Population Size",
       y = "Frequency") +
  theme_classic() + 
 facet_grid(model ~ .)
```

```{r calculate aggregation metrics}

```
