---
title: "Willoughby ectoparasite model"
author: "Anna Willoughby and John Drake"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=FALSE)
library('deSolve') # solving ordinary differential equations
library(dplyr) # data wrangling
set.seed(1993)
```

## Model

In this document we seek to use a birth-death-immigration process to model the number of ectoparasites on a host.
A full description of the model is at https://github.com/DrakeLab/willoughby-grooming-model.

We start by writing functions for the birth rate, death rate, and immigration rate.
In all cases, we let the state variable ($x$) be the first parameter.
We also set some default parameters to make things simpler later.

```{r, functions}
birthrate <- function(x, b0=1, theta=1) b0*x*(x/(theta+x)) # function for birth rate with Allee effect (theta)
deathrate <- function(x, d0=0, d1=0.01) (d0+d1*x)*x # function for death rate
immigrationrate <- function (x, iota=1) iota # function for immigration rate
```

It might be nice to know what the maximum population size is.
This would require doing a little algebra and it may require numerical methods.

Now, we study the deterministic model numerically, as this will help to choose parameter values.

```{r, deterministic model}
# define the function for population growth rate
f <- function(t, x, parms){
  dx <- birthrate(x, parms[1], parms[2]) - 
    deathrate(x, parms[3], parms[4]) + 
    immigrationrate (x, parms[5])
  
  return(list(dx))
}

parms <- c(b0=1, theta=1, d0=0, d1=0.01, iota=1) # set initial conditions parameters

solution <- ode(y=1, times=seq(1:100), func=f, parms=parms)  # set initial conditions: start with one individual

plot(solution, main='Deterministic trajectory')

```

### solve for the carrying capacity

```{r max pop calculation}
# when does f = 0 ? 

# Define the equation to solve # [can prop simplify with f above....]
equilibrium_eq <- function(x, parms) {
  with(as.list(parms), {
    birthrate <- b0*x*(x/(theta+x))
    deathrate <- (d0 + d1*x) * x
    immigrationrate <- iota
    return(birthrate - deathrate + immigrationrate)
  })
}

# Solve for equilibrium
parms <- c(b0=1, theta=1, d0=0, d1=0.01, iota=1)
root <- uniroot(equilibrium_eq, c(0, 101), parms=parms)$root # max population size is slightly >100 (need to make sure the interval includes a solution)

print(paste("Maximum population size:", root))
```


## Stochastic Model

Now, we want to simulate a birth-death-immigration process.
We start with some hyperparameters of the simulation.

```{r set some simulation rules}
num_simulations <- 100  # Number of times to run the simulation
# max.events <- 1000 # maximum events
max.time <- 200 # 2 year max for a squirrel  
```

Here we simulate the process

```{r set up simulation}
# Store results in a list
all_simulations <- vector("list", num_simulations)

for (sim in 1:num_simulations) {

events <- 0 #initialize the total number of events
time <- 0  #initialize the current time
x <- 0 #initialize population size

# Create a dataframe to store results
simulation_results <- data.frame(event = numeric(),
                                 time = numeric(),
                                 population = numeric(), 
                                 event_type = character(),
                                 sim_id = numeric())
# no longer doing event length (events <= max.events)
while(tail(time,1) <= max.time ){
  
  # some bookkeeping
  events <- events + 1 #increment the number of events
  current.size <- x[events] #set local current population size
  current.time <- time[events]
  
  # get the rates and store locally
  b <- birthrate(x=current.size)
  d <- deathrate(x=current.size)
  i <- immigrationrate(x=current.size)
  
  # find the next event time
  total.rate <- b+d+i
  
  increment.time <- -log(1-runif(1))/total.rate # exponential time to next event
  
  # find next event type John's way
  # change <- sample(c(-1, 1), size=1, prob = c(d/total.rate,(b+i)/total.rate))
  
  # Determine event type Anna's way
  event_probs <- c(d, b, i) / total.rate # what is the rate right now for each of the event types? 
  event_outcome <- sample(c("death", "birth", "immigration"), size = 1, prob = event_probs) # sample what happens 
  change <- ifelse(event_outcome == "death", -1, 1) # what will be the population change? 
  
  # Store results in dataframe
  simulation_results <- rbind(simulation_results, data.frame(event = events,
                                                             time = current.time + increment.time,
                                                             population = current.size + change,
                                                             event_type = event_outcome,
                                                             sim_id = sim))
  
  # update state variables and store
  time <- c(time, current.time+increment.time)
  x <- c(x,current.size+change)
}

# Store this simulation's results
  all_simulations[[sim]] <- simulation_results
}
all_simulations <- saveRDS(all_simulations, "data/all_simulations.Rdata")
# Combine all simulation results into one dataframe
# final_results <- do.call(rbind, all_simulations)

# plot(time,x, type = 'b', xlab='Time', ylab='Ectoparasite population size') # this is for single sim 

library(ggplot2)
# make a plot with all the simulations 
# ggplot(final_results, aes(x = time, y = population, group = sim_id, color = factor(sim_id))) +
#  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
#  labs(title = "Ectoparasite Population Over Time (*n* = 100 simulations)", 
#       x = "Time", y = "Population Size", color = "Simulation ID") +
#  theme_classic() +
#  theme(legend.position = "none")  # hide legend cause too many lines
```

Now, I need a way to sample from these data.

```{r use the event dataframe to calculate values}
## FLEA POPULATION TRAITS 

# value for one simulation
# flea_total_pop =  sum(simulation_results$event_type == "immigration") + sum(simulation_results$event_type == "birth")

# how many total fleas across all simulations
flea_total_pop_list <- sapply(all_simulations, function(sim_df) {
  sum(sim_df$event_type == "immigration") + sum(sim_df$event_type == "birth")
})
hist(flea_total_pop_list)

# need to test if any flea pops go extinct

# crowding

## HOST TRAITS ## need to fix time units --- currently in flea reproduction units..... 
# calculate mean daily end in flea pop -- > this is the intensity! 
## round time to the nearest integer and keep only the last observation per time unit
end_of_day_flea_pop <- simulation_results %>%
  mutate(time_unit = floor(time)) %>%  # Convert time to integer units
  group_by(time_unit) %>%
  slice_tail(n = 1) %>%  # Get the last observation per unit
  ungroup() %>%
  select(time_unit, population)  # Keep relevant columns
## calculate based on sum of the daily pops / end time duration 
mean_daily_end <- sum(end_of_day_flea_pop$population)/ end_of_day_flea_pop[nrow(end_of_day_flea_pop), 1]
mean_daily_end 
```



```{r, sampling-function}
sample_population <- function(event_times, population_sizes, sample_times, initial_pop = NA) {

    # Ensure event_times and population_sizes are vectors of the same length
  if (length(event_times) != length(population_sizes)) {
    stop("event_times and population_sizes must be vectors of the same length.")
  }
  
  # Ensure event_times are sorted in increasing order
  if (is.unsorted(event_times)) {
    stop("event_times must be sorted.")
  }
  
  # Find the interval indices for each sampling time
  indices <- findInterval(sample_times, event_times)
  
  # Return the value of x from the sample intervals
  result <- population_sizes[indices]

  return(result)
}
```

Here I demonstrate use of this function.

```{r, sample-histogram}
s <- sample_population(time, x, seq(0.1,8, by=0.01)) 
hist(s, main='Histogram of ectoparasite burden', xlab='Individuals')
```

Now, let's create a version of the model that runs until the age limit of a host.
First we have to figure out how long a host (rock squirrel) lives in units of flea reproduction.
We will suppose a rock squirrel lives for five years and that a flea reproduces 100 off spring in one year.
That means that a rock squirrel lives through 500 reproduction units of fleas.
If the *per capita* flea birth rate is 1.0, then on average a rock squirrel lives to time 500.
(Note: this could be slow to compute.)

```{r max age of host in flea reproduction times}
max.time <- 500 #later change to 500
max.events <- 1000
```

Here we simulate the process

```{r write the simulation}
events <- 0 #initialize the total number of events
time <- 0  #initialize the current time
x <- 0 #initialize population size

while(tail(time,1) <= max.time ){
  
  # some bookkeeping
  events <- events + 1 #increment the number of events
  current.size <- tail(x,1)  # Always take the last value
  current.time <- tail(time,1) # Always take the last value
  
  # get the rates and store locally
  b <- birthrate(x=current.size)
  d <- deathrate(x=current.size)
  i <- immigrationrate(x=current.size)
  
  # find the next event time
  total.rate <- b+d+i
  
  increment.time <- -log(1-runif(1))/total.rate # exponential time to next event
  
  # find next event type
  change <- sample(c(-1, 1), size=1, prob = c(d/total.rate,(b+i)/total.rate))
  
  # update states and store
  time <- c(time, current.time+increment.time)
  x <- c(x,current.size+change)
}

plot(time,x, type = 'b', xlab='Time', ylab='Ectoparasite population size')
```

Here we write the simulation loop in a function that allows us to simulate the trajectory of infection in an individual of age $A$.

We will also generalize the function to arbitrary input parameters so that we don't rely on the defaults.

```{r}
simulate <- function(A = 500,
                     parms = NULL,
                     plot = FALSE) {
  events <- 0 #initialize the total number of events
  time <- 0  #initialize the current time
  x <- 0 #initialize population size
  
  if (is.null(parms)) {
    b0 <- 1
    theta <- 1
    d0 <- 0
    d1 <- 0.01
    iota <- 1
    
  } else {
    
    b0 <- parms$b0
    theta <- parms$theta
    d0 <- parms$d0
    d1 <- parms$d1
    iota <- parms$iota
    
  }
  
  while (tail(time, 1) <= A) {
    # some bookkeeping
    events <- events + 1 #increment the number of events
    current.size <- x[events] #set local current population size
    current.time <- time[events]
    
    # get the rates and store locally
    b <- birthrate(x = current.size, b0=b0, theta=theta)
    d <- deathrate(x = current.size, d0=d0, d1=d1)
    i <- immigrationrate(x = current.size, iota=iota)
    
    # find the next event time
    total.rate <- b + d + i
    
    increment.time <-
      -log(1 - runif(1)) / total.rate # exponential time to next event
    
    # find next event type
    change <-
      sample(c(-1, 1),
             size = 1,
             prob = c(d / total.rate, (b + i) / total.rate))
    
    # update states and store
    time <- c(time, current.time + increment.time)
    x <- c(x, current.size + change)
  }
  
  if (plot == T) {
    plot(time,
         x,
         type = 'b',
         xlab = 'Time',
         ylab = 'Ectoparasite population size')
  }
  
  return(data.frame(time, x))
  
}
```
Here we apply the function to an individual that lives 100 time units (one year)

```{r}
out1 <- simulate(A=100, plot=TRUE)
```

Here we apply the function to a different set of parameters (much smaller immigration rate).

```{r}
parameters <- list(b0=1, theta=1, d0=0, d1=0.01, iota=0.02)
out2 <- simulate(A=100, parms = parameters, plot=TRUE)
```


```{r calculate traditional infection metrics from host perspective}
# calculate prevalence

# mean abundance

# mean intensity


```


