---
title: "Workflow-v2"
author: "Anna Willoughby"
date: "2025-05-22"
output: pdf_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library('deSolve') # solving ordinary differential equations
library(dplyr) # data wrangling
library(ggplot2) # pretty plots
library(lhs) # latin hypercube sampling
library(tidyr) # pipes
library(parallel) # for running in parallel 
library(gridExtra) # plotting multipanels 
set.seed(1993)

```

## Model

In this document we seek to use a birth-death-immigration process to model the number of ectoparasites on a host. A full description of the model is at https://github.com/DrakeLab/willoughby-grooming-model.

We start by writing functions for the birth rate, death rate, and immigration rate.
In all cases, we let the state variable ($x$) be the first parameter.
We also set some default parameters to make things simpler later.


```{r, functions}
# function for birth rate with Allee effect (theta)
birthrate <- function(x, b0, theta, K){
  b <- b0*x*(x/(theta+x)) # baseline births with Allee
  b_dd <- b*x*(1 - x/ K) # birth decreases at high densities
  b_wA_dd <- pmax(0, b_dd)
  b_wA_dd # birth rate with Allee and density dependence
}

# function for immigration rate (iota)
# baseline iota, and resource linked increase
immigrationrate <- function (x, iota_baseline, iota_r, scenario){ 
  iota = iota_baseline*(1 + scenario * iota_r)   
  iota
}

# function for programmed grooming 
# mid-grooming at a baseline level when resources are at 50 
groomrate_prg <- function(pop.size, groom_prg_bl, scenario_id){
  death_rate <- pmin(pop.size, groom_prg_bl * scenario_id / 50)
  death_rate
}

# function for stimulus grooming deathrate based on pop size 
# mid-grooming at a baseline level when resources are at 50 
groomrate_stm <- function(pop.size, groom_stim_bl, scenario_id){
  death_rate <- pop.size*( (groom_stim_bl * (scenario_id / 50)) /100)
  death_rate
}

# Function to get cumulative groom rate from programmed and stimulus grooming 
get_groom_rate <- function(pop.size, groom_prg_bl, groom_stim_bl, scenario_id) {
  prg_rate = groomrate_prg(pop.size, groom_prg_bl, scenario_id)
  stm_rate = groomrate_stm(pop.size, groom_stim_bl, scenario_id) 
  groom_rate =  pmin(pop.size, prg_rate + stm_rate)
  groom_rate
}


# Function for density dependent deaths 
deathrate_dd <- function(pop.size, dmax = 1, x_mid = 50, s = 20) {
     dmax / (1 + exp(-(pop.size - x_mid) / s))
}

# set initial conditions parameters

## birth rate is standard across all scenarios, right now as 37.5 flea births per year (0. per day)
## we are going to vary parameters for:  
# theta_vals <- seq(0, 10, by = 1) # 10 different values

# iota_bl_vals <- seq(0, 0.5, by = 0.1) # Keep for all the same 
# iota_r_vals <- seq(0, 0.5, by = 0.05) # 10 different values
scenario_vals <- seq(1,100, by = 1) # 100 different values (no grooming is an explicit scenario) 
groom_prg_bl_vals <- c(0,20) # grooming baseline [two scenarios: no programmed or baseline of 20]
groom_stim_bl_vals <- c(0, 20) # grooming baseline  [two scenarios: no stimulus or baseline of 20]

# construct full matrix of theta and iota options 
ti_axes <- expand.grid(
  #theta = theta_vals, # for now doing one theta value
  #iota_baseline = iota_bl_vals,  # for now doing one iota_baseline value
  #iota_r = iota_r_vals,  # for now doing one iota resource link value
  scenario = scenario_vals, 
  groom_prg_baseline = groom_prg_bl_vals, 
  groom_stm_baseline = groom_stim_bl_vals 
)

# Construct parameter set
param_grid <- ti_axes
param_grid$b0 <- log(78.75)/89 # constant [log(female fecundity per lifetime)/generation time]
param_grid$initial_pop <- 1
param_grid$theta <- 8
param_grid$iota_baseline <- 0.5
param_grid$iota_r <- 0.2
```


## Stochastic Model

Now, we want to simulate a birth-death-immigration process.
We start with some hyperparameters of the simulation.

```{r set some simulation rules}
# Individual Experiment 
num_simulations <- 100  # Number of times to run the simulation
# max.events <- 1000 # maximum events
max.time <- 730 # 2 year max for a squirrel  

```

Here we simulate the process

```{r run stochastic models}

# Define the task grid: 
task_grid <- expand.grid(
    param_id = 1:nrow(param_grid),
    sim_id  = 1:num_simulations)
 
# Define running the whole model as a function 
run_one_parameter_set <- function(task_i){

    param_id <- task_grid[task_i, 1]
    sim_id   <- task_grid[task_i, 2]
    
    message("Parameter set ", param_id, ", Simulation", sim_id)
    
    parms <- param_grid[param_id, ]

    events <- 0
    time <- 0
    x <- parms$initial_pop

    out_list <- list()

    while (time[length(time)] <= max.time) {

      events <- events + 1
      current.size <- x[length(x)]
      current.time <- time[length(time)]

      b <- birthrate(current.size, parms$b0, parms$theta, 50) # 50 as K
      g <- get_groom_rate(current.size, 
                   parms$groom_prg_baseline,
                   parms$groom_stm_baseline, 
                   parms$scenario)
      #message("Current size", current.size, ", groom rate", g) # mid function checks
      d <- deathrate_dd(current.size) 
      iota <- immigrationrate(current.size, 
                    parms$iota_baseline, parms$iota_r, 
                    parms$scenario)

      total.rate <- b + g + d + iota

      if (is.na(total.rate) || total.rate <= 0) break

      increment.time <- -log(1 - runif(1)) / total.rate

      event_probs <- c(d, g, b, iota) / total.rate
      event_outcome <- sample(
        c("death","groomed","birth", "immigration"),
        size = 1,
        prob = event_probs
      )

      change <- ifelse(event_outcome %in% c("death", "groomed"), -1, 1)

      new_size <- max(0, current.size + change)
      new_time <- current.time + increment.time

      out_list[[events]] <- data.frame(
        event = events,
        time = new_time,
        population = new_size,
        event_type = event_outcome,
        sim_id = sim_id,
        param_id = param_id
      )

      time <- c(time, new_time)
      x <- c(x, new_size)
    }

    if (length(out_list) == 0) {
      data.frame(
        event = numeric(),
        time = numeric(),
        population = numeric(),
        event_type = character(),
        sim_id = numeric(),
        param_id = numeric()
      )
    } else {
      do.call(rbind, out_list)
    }}

# Now run for the full simulation across the 10 clusters 

  
# Now run for the full simulation across the 10 clusters 
system.time({

  cl <- makeCluster(10)
  
  # Move libraries to the clusters 
  clusterEvalQ(cl, library(tidyverse))

  # Move main environment items to each of the clusters
  clusterExport(cl, 
                c("param_grid","max.time", "task_grid", 
                   "birthrate", "deathrate_dd", "immigrationrate",
                  "get_groom_rate", "groomrate_prg", "groomrate_stm"
                  ))
  
  simulation <- parLapply(cl, 1:nrow(task_grid), run_one_parameter_set)

  on.exit(stopCluster(cl), add = TRUE)
})

saveRDS(simulation, "data/theta_rig_links/sim_t7_rig_bi02_br04.Rdata")
# rename param grid
output_params <- param_grid
# create id col 
output_params$param_id = 1:nrow(output_params)
# merge task grid and params 
output_params <- left_join(task_grid, output_params, by = "param_id")
# save the task_grid 
write.csv(task_grid, file = "data/theta_rig_links/task_grid_t7_rig.csv")

```
```{r plot simulation time series for midgrooming scenarios}
# Plot example time series 
# Load simulation if needed
# system.time({
# simulation <- readRDS("data/sim_100gs_10sims.Rdata")
# })

# Create param id on the parameters df for linking to each simulation 
param_grid$param_id = 1:nrow(param_grid)

system.time({
sim_unlisted <- data.table::rbindlist(simulation)
})

# Link parameters to each simulation 
system.time({
sim_unlisted_wp <- left_join(sim_unlisted, param_grid, by = "param_id")
})

# resave levels 
sim_unlisted_wp$groom_strategy <- factor(sim_unlisted_wp$groom_strategy, levels = 
                                           c("none", "programmed", "stimulus"))

# Plot simulation over a lifetime 
sim_lt_plot <- ggplot(sim_unlisted_wp[sim_unlisted_wp$scenario == 20,],
                       aes(x = time, y = population, color = groom_strategy,
                           group = interaction(param_id, sim_id))) +
  geom_line(alpha = 0.75, linewidth = 0.8) +
  #geom_hline(yintercept = mean(sim754$theta), 
 #            linetype = "dashed", color = "grey40", linewidth = 0.5) +
  facet_wrap(~groom_strategy, ncol = 1, scales = "free") +
  #scale_color_manual(values = strategy_colors) +
  labs(
    title    = "Flea population dynamics by grooming strategy",
    x = "Day", y = "Ectoparasite population size", color = "Grooming Strategy"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 12))
# check all good 
sim_lt_plot 

# Save plot 
png(filename="figures/Figure3_lt_20gs_rig_t6.png", width = 16, height =8, unit = 'in', res = 300)
sim_lt_plot
dev.off()



```



Now, I need a way to sample from these data.

```{r calculate traits - total pop}
# read in simulation data 
# sim10 <- readRDS("data/simulation_10rep.Rdata")
# sim <- readRDS("data/simulation.Rdata")
# sim100 <- readRDS("data/simulation100.Rdata")


# optional flatten to dataframe
# results_sim10 <-  do.call(rbind, sim10)
# results_sim10$id <- paste(results_sim10$param_id, "-", results_sim10$sim_id, sep = "")

# create functions to calculate output values 

# FLEA POPULATION TRAITS - calculate relevant traits 
## Snapshot/Sampling Event: the count of fleas at one snapshot (e.g., end of day)

# function to calculate daily snapshot 

end_of_day_flea_pop <- function(sim){
  
  init_row <- data.frame(  # pre load initial conditions simulations 
    event = 0,
    time = 0,
    population = 1,
    event_type = "initial",
    sim_id = sim$sim_id[1],
    param_id = sim$param_id[1]
  )

  sim %>% # Apply operations to the dataset
    bind_rows(init_row, .) %>%
    mutate(time_unit = floor(time)) %>% # Convert time to integer units
    group_by(time_unit) %>%
    slice_tail(n = 1) %>% # Get the last observation per unit
    ungroup() %>%
    select(time_unit, population) %>% # Keep relevant columns
    
    # create full sequence
    tidyr::complete(time_unit = 0:780) %>%
    
    # carry previous value forward
    tidyr::fill(population, .direction = "down") %>% 
    dplyr::filter(time_unit != 0) %>% # get rid of the initial seeding
    pull(population) # only get a vector of the population sizes 
}

## values for one simulation

summarize_simulations <- function(simulation_list) {
  
  result <- lapply(seq_along(simulation_list), function(i) {
    
    eod_burdens <- end_of_day_flea_pop(simulation_list[[i]])
    
     eod_burden_list <-  list( # save list of the eod values for each sim 
       sim_id = i,
       eod_burdens = eod_burdens)


    sim_df <- simulation_list[[i]]
    
    
    data.frame(
      list_id = i,
      sim_id = unique(sim_df$sim_id)[1],
      param_id = unique(sim_df$param_id)[1],
      ind_metric_total_pop = sum(sim_df$event_type == "immigration") + # calculate total pop.size
                        sum(sim_df$event_type == "birth"),
      ind_metric_total_groom = sum(sim_df$event_type == "groomed"), # calculate total flea groomed
      
      ind_metric_infest_days = sum(eod_burdens  > 0),
      ind_metric_mean_daily_pop = mean(eod_burdens)
    )
  })
  
  do.call(rbind, result)
}

# run for each simulation 
system.time({
  flea_stat_df <- summarize_simulations(simulation) # each simulation
}) 

# summarize by parameters 
flea_stat_per_param <- flea_stat_df %>% 
  group_by(param_id, sim_id) %>% 
  summarise(ind_metric_total_pop = mean(ind_metric_total_pop), 
            ind_metric_total_grm = mean(ind_metric_total_groom), 
            ind_metric_inf_days = mean(ind_metric_infest_days), 
            ind_metric_avg_dy_burden = mean(ind_metric_mean_daily_pop))

# merge in parameters 
param_grid$param_id <-  seq_along(tig_axes[,1])
flea_stat_per_param <- left_join(flea_stat_per_param, param_grid, by ="param_id")

# Save output 
write.csv(flea_stat_per_param, file ="data/exp1_output_gs_ri.csv")

# If I need to read in the saved data 
# flea_stat_per_param <- read.csv(file = "data/exp1_output.csv")
# lets look at no grooming for now: 

# now for a mid grooming scenarios (50) plot, faceted by different strategies 

# plot individual infestation traits by resource 
## convert into long data 

flea_stat_per_param_long <- flea_stat_per_param %>%
  pivot_longer(cols = starts_with("ind_metric_"), 
                names_to = "metric",
                values_to = "value") 

# confirm groom strategy factors 
flea_stat_per_param_long$groom_strategy <- factor(flea_stat_per_param_long$groom_strategy,
        levels = c("none", "programmed", "stimulus"))

# K as the sim value 
flea_stat_per_param_k <- flea_stat_per_param_long %>%
  dplyr::filter(metric == "ind_metric_avg_dy_burden") 

flea_stat_per_paramsim_k <- flea_stat_per_param_k %>% 
  group_by(param_id, groom_strategy, scenario) %>%
  summarise(m = mean(value),
            v = var(value))

flea_stat_per_paramsim_k$k <- (flea_stat_per_paramsim_k$m)^2 / 
  (flea_stat_per_paramsim_k$v - flea_stat_per_paramsim_k$m)

ggplot(flea_stat_per_paramsim_k, aes(x = scenario, y = k)) + 
  geom_line() +
  facet_wrap(~groom_strategy)
# Now Resources jointly impacts immigration and grooming 


## convert so resources affects immigration same as grooming 
# calculate appropriate 

# confirm grooming strategy levels
flea_stat_per_param_long_ri_link$groom_strategy <- factor(flea_stat_per_param_long_ri_link$groom_strategy,
                                                          levels = c("none", 
                                                                     "programmed", 
                                                                     "stimulus")) 
# prepare individual metric plots 

# PLOT OF RESOURCE on individual metrics with immigration-grooming paired 
## total population 
totpop_plot <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$metric == "ind_metric_total_pop",], 
                      aes(x = scenario, group = groom_strategy, y = value, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "total population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
totpop_plot

# remove legend for multi-panel plot: 
totpop_plot_nl <- totpop_plot +
  theme(legend.position = "none",
        axis.title.x = element_blank()) # use joint label instead

## average eod burden 
dayburden_plot <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$metric == 
                                                            "ind_metric_avg_dy_burden",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 3) +
  labs(x = "resource availability",
       y = "mean daily burden") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
dayburden_plot

# remove legend for multi-panel plot: 
dayburden_plot_nl <- dayburden_plot +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) # use joint label instead

## total fleas groomed  
grooming_effort_plot <- ggplot(flea_stat_per_param_long[                    flea_stat_per_param_long$metric == 
                                                            "ind_metric_total_grm",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "total groomed population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
grooming_effort_plot

# remove legend for multi-panel plot: 
grooming_effort_plot_nl <- grooming_effort_plot +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none")
  
## infestation days 
inf_days_plot <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$metric == 
                                                            "ind_metric_inf_days",], 
                      aes(x = scenario, y = round(value/780*100,2), 
                          group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "days with an infestation (%)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank(),
         legend.position = "none") # use joint label instead
inf_days_plot # check all good 

# Multipanel plot for one theta value (2)
png(filename="figures/FigureX_ind_metrics_rig_t6.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(totpop_plot_nl, dayburden_plot_nl , grooming_effort_plot_nl, inf_days_plot, 
             nrow = 4,
             bottom = "resource availability")
dev.off()

# Plot summary stats of this: 
# -- Plot 2: Strategy means + IQR ribbons --
flea_stat_df_sum <- left_join(flea_stat_df, param_grid, by = "param_id")
summary_df <- flea_stat_df_sum |>
  group_by(groom_strategy, param_id, scenario) |>
  summarise( mean_daily_burden = mean(ind_metric_mean_daily_pop),
            lo     = quantile(ind_metric_mean_daily_pop, 0.25),
            hi     = quantile(ind_metric_mean_daily_pop, 0.75),
            .groups = "drop")

summary_mb <- ggplot(summary_df, 
                     aes(x = scenario, color = groom_strategy, fill = groom_strategy)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, color = groom_strategy), alpha = 0.15) +
  geom_line(aes(y = mean_daily_burden), linewidth = 1.1) +
  labs(
    title    = "Mean daily flea load ± IQR by grooming strategy",
    x = "resources", y = "Mean daily flea count", color = "Strategy", fill = "Strategy"
  ) +
  theme_minimal(base_size = 13)
summary_mb 


```

```{r subsetting the eod }

# create a function to sample only a portion of the days (e.g. last 25%)
subset_simulation <- function(sim_df, range_percent) {
  # Ensure range_percent is a valid numeric vector of length 2 (e.g., c(0, 10) or c(75, 100))
  if (length(range_percent) != 2 || any(range_percent < 0) || any(range_percent > 100)) {
    stop("range_percent must be a numeric vector of length 2 between 0 and 100.")
  }
  
  # Get the start and end time from the simulation
  min_time <- min(sim_df$time_unit)
  max_time <- max(sim_df$time_unit)

  # Compute the time range corresponding to the percentage range
  time_start <- min_time + (range_percent[1] / 100) * (max_time - min_time)
  time_end <- min_time + (range_percent[2] / 100) * (max_time - min_time)

  # Subset the dataframe based on the calculated time range
  sim_df %>%
    filter(time_unit >= time_start & time_unit <= time_end)
}

# limit population snapshots to latter 75% (eliminating burn-in period)
# sim_eod_pop_last75 <- lapply(eod_burden_list, subset_simulation, range_percent = c(26, 100))

# Create master dataframe (likely not needed)
# sim_eod_pop_last75_combined <- bind_rows(
#  lapply(seq_along(sim_eod_pop_last75), function(i) {
#    sim_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
#  })
# )
```

```{r population sim}
# Population experiment 
# overwrite for population
num_simulations <- 1  # Number of times to run the simulation - this time just once very long
max.time <- 3650 # 10 years -- > beyond on squirrel lifetime 
# Define the task grid: 
task_grid <- expand.grid(
    param_id = 1:nrow(param_grid), # using same parameters
    sim_id   = 1:num_simulations)

# run the simulation for the population 
system.time({

  cl <- makeCluster(10)
  
  # Move libraries to the clusters 
  clusterEvalQ(cl, library(tidyverse))

  # Move main environment items to each of the clusters
  clusterExport(cl, 
                c("groom_prg_scenarios","groom_stm_scenarios", "groom_cst_scenarios",
                   "param_grid","max.time", "task_grid", 
                   "birthrate", "deathrate_dd", "immigrationrate", 
                  "groom_strategies", "get_groom_rate", "groomrate_prg", "groomrate_stm", "groomrate_cst"
                  ))
  
  simulation_pop <- parLapply(cl, 1:nrow(task_grid), run_one_parameter_set)

  on.exit(stopCluster(cl), add = TRUE)
})

# save output 
saveRDS(simulation_pop, "data/simulation_pop.Rdata")
```


```{r plot simulation time series for midgrooming scenarios}
# Plot example time series 
# Load simulation if needed
# system.time({
# sim_pop <- readRDS("data/simulation_pop.Rdata")
# })

# Create param id on the parameters df for linking to each simulation 
param_grid$param_id = 1:nrow(param_grid)

system.time({
sim_pop_unlisted <- data.table::rbindlist(sim_pop)
})


# Link parameters to each simulation 
system.time({
sim_pop_unlisted_wp <- left_join(sim_pop_unlisted, param_grid, by = "param_id")
})

# resave levels 
sim_pop_unlisted_wp$groom_strategy <- factor(sim_pop_unlisted_wp$groom_strategy, levels = 
                                           c("none", "programmed", "stimulus", "costly"))

# Plot simulation of population
sim_pop_ex_plot <- ggplot(sim_pop_unlisted_wp[sim_pop_unlisted_wp$theta == 9 &
                       sim_pop_unlisted_wp$scenario == 50 & 
                       sim_pop_unlisted_wp$iota == 0.5 &  
                       sim_pop_unlisted_wp$groom_strategy != "costly",],
                       aes(x = time, y = population, color = groom_strategy,
                           group = interaction(param_id, sim_id))) +
  geom_line(alpha = 0.75, linewidth = 0.8) +
  #geom_hline(yintercept = mean(sim754$theta), 
 #            linetype = "dashed", color = "grey40", linewidth = 0.5) +
  facet_wrap(~groom_strategy, ncol = 1, scales = "free") +
  #scale_color_manual(values = strategy_colors) +
  labs(
    title    = "Flea population dynamics by grooming strategy",
    x = "Day", y = "Ectoparasite population size", color = "Grooming Strategy"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 12))
# check all good 
sim_pop_ex_plot 
# Save plot 
png(filename="figures/FigureX_pop_gs_i05_t4.png", width = 16, height =8, unit = 'in', res = 300)
sim_pop_ex_plot
dev.off()

```

```{r calculate host population traits}
# calculate the autocorrelation and the needed space between pops 
# find_acf_gap <- function(x, cutoff = 0.1, lag.max = 200) {
#  a <- acf(x, plot = FALSE, lag.max = lag.max, na.action = na.pass)
#  lags <- a$lag[-1]
#  acfs <- a$acf[-1]
  
#  idx <- which(abs(acfs) < cutoff)[1]
  
#  if (is.na(idx)) return(NA)
#  lags[idx]
# }

# find_acf_gap(eod_burdens_list[[1]]$population)

# Calculate simulation values 
# create a list of time points to get the values 
sample_days <- seq(949, 3449, by = 25) # last 75% of 10 years, every 25 days apart 

# update eod function for full simulation: 
end_of_day_flea_pop_3650 <- function(sim){
  
  init_row <- data.frame(  # pre load initial conditions simulations 
    event = 0,
    time = 0,
    population = 1,
    event_type = "initial",
    sim_id = sim$sim_id[1],
    param_id = sim$param_id[1]
  )

  sim %>% # Apply operations to the dataset
    bind_rows(init_row, .) %>%
    mutate(time_unit = floor(time)) %>% # Convert time to integer units
    group_by(time_unit) %>%
    slice_tail(n = 1) %>% # Get the last observation per unit
    ungroup() %>%
    select(time_unit, population) %>% # Keep relevant columns
    
    # create full sequence
    tidyr::complete(time_unit = 1:3650) %>%
    
    # carry previous value forward
    tidyr::fill(population, .direction = "down") %>% 
    dplyr::filter(time_unit >= 1, time_unit <= 3650) %>% # get rid of the initial seeding
    pull(population) # only get a vector of the population sizes 
}


## function to sample across the simulation
sample_snapshots_one_pop <- function(simulation_list, sample_days, i) {
  
    pop_eod_burdens <- end_of_day_flea_pop_3650(simulation_list[[i]])
    
    pop_eod_burden_list <-  list( # save list of the eod values for each sim 
       sim_id = i,
       eod_burdens = pop_eod_burdens)
    
    pop_eod_burdens_df <- data.frame(time_unit = c(1:3650), eod_burden = pop_eod_burdens)
    
    pop_eod_burden_last75 <- subset_simulation(pop_eod_burdens_df,
                                               range_percent = c(26, 100))
    pop_sampled_burden <- pop_eod_burden_last75[pop_eod_burden_last75$time_unit %in% sample_days, ]
    pop_sampled_burden
}

# load sim_pop if needed 
# simulation_pop <- load("data/simulation_pop.Rdata")
# create a list of the sampled day values per pop sim 
pop_sampled_burden_list <- lapply(
  seq_along(simulation_pop),
  function(i) sample_snapshots_one_pop(simulation_pop, sample_days, i)
)

# Now summarize for these 100 samples 
summarize_pop_simulations <- function(simulation_list) {
  
  result <- lapply(seq_along(simulation_list), function(i) {
    
    # save output as a dataframe 
    
    pop_sim_df <- simulation_list[[i]]
    
    pop_metric_prev <- sum(pop_sim_df$eod_burden > 0)
    pop_metric_abun <- sum(pop_sim_df$eod_burden) / 100
    pop_metric_inte <- pop_metric_abun / pop_metric_prev
      
    mu <- mean(pop_sim_df$eod_burden)
    v <- var(pop_sim_df$eod_burden)
    k <- mu^2 / (v - mu)
      
    pop_metric_crowd <- mu + mu / k
    pop_metric_patch <- 1 + 1 / k
    
    data.frame(
      list_id = i,
      param_id = i,
      pop_metric_prev = pop_metric_prev,
      pop_metric_abun = pop_metric_abun,
      pop_metric_inte = pop_metric_inte,
      mu = mu,
      v = v,
      pop_metric_k = k,
      pop_metric_crowd = pop_metric_crowd,
      pop_metric_patch = pop_metric_patch
    )
  }) 
  
  do.call(rbind, result)
}

# run for each simulation 
system.time({
  pop_stat_df <- summarize_pop_simulations(pop_sampled_burden_list) # each simulation
}) 

```

```{r plot population measures across resource scenarios }
# merge in parameters 
pop_stat_per_param <- left_join(pop_stat_df, param_grid, by ="param_id")

# save the pop eod spaced burdens
saveRDS(pop_stat_per_param, "data/sim_pop_metrics.Rdata")

# If I need to load the saved pop_stat_per_param 
# pop_stat_per_param <- readRDS("data/sim_pop_metrics.Rdata")

# correct abundance to mean abundance 
pop_stat_per_param$pop_metric_abun <- pop_stat_per_param$pop_metric_abun / 100

# lets look at no grooming for now: 
## filter to the no grooming
pop_stat_per_param_ng <- pop_stat_per_param %>% 
  dplyr::filter(groom_strategy=="none")
## plot heatmap of prevalence
prev_it_ng_plot <- ggplot(pop_stat_per_param_ng, 
       aes(x = theta, y = iota, fill = pop_metric_prev)) +
  geom_tile() +
  scale_y_continuous(breaks = seq(0, 0.5, by = 0.1)) +
  scale_fill_gradientn(colours = c("grey80", viridis::viridis(255, direction = -1)),
                       values = scales::rescale(c(1, 101)),
                       name = "population prevalence", 
                       breaks = c(0,25,50,75,100),
                       labels = c("1","25","50","75","100")) +
  labs(x = "theta",
       y = "iota",
       fill = "prevalence") +
  theme_minimal()
prev_it_ng_plot # check all good

png(filename="figures/FigureX_ti_ng_prev.png", width = 16, height =8, unit = 'in', res = 300)
prev_it_ng_plot
dev.off()

# now for a mid grooming scenarios (50) plot, faceted by different strategies 
pop_stat_per_param_g50 <- pop_stat_per_param %>% 
  dplyr::filter(scenario==50)
## define levels of groom_stragety 
pop_stat_per_param_g50$groom_strategy <- factor(pop_stat_per_param_g50$groom_strategy, 
                                                 levels = c("none", "programmed", "stimulus",
                                                            "costly"))
               
## plot heatmap of prevalence
ti_midg_plot <- ggplot(pop_stat_per_param_g50, 
       aes(x = theta, y = iota, fill = pop_metric_prev)) +
  geom_tile() +
  facet_wrap(~groom_strategy, ncol = 1) +
  scale_y_continuous(breaks = seq(0, 0.5, by = 0.1)) +
  scale_fill_gradientn(colours = c("grey80", viridis::viridis(255, direction = -1)),
                       values = scales::rescale(c(1, 101)),
                       name = "prevalence", 
                       breaks = c(0,25,500,75,100),
                       labels = c("0","25","50","75","100")) +
  labs(x = "theta",
       y = "iota",
       fill = "prevalence") +
  theme_minimal()

ti_midg_plot # check all good 

png(filename="figures/FigureX_ti_midg_prev.png", width = 16, height =8, unit = 'in', res = 300)
ti_midg_plot
dev.off()

# plot individual infestation traits by resource 
## convert into long data 
pop_stat_per_param_long <- pop_stat_per_param %>%
  pivot_longer(cols = starts_with("pop_metric_"), 
                names_to = "metric",
                values_to = "value") 

# confirm groom strategy factors 
pop_stat_per_param_long$groom_strategy <- factor(pop_stat_per_param_long$groom_strategy,
                                                  levels = c("none", 
                                                             "programmed", 
                                                             "stimulus",
                                                             "costly"))

# plot individual metrics for only resource-induced changes to grooming, and no immigration 

pop_met_rg_ni_plot <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0, ], 
       aes(x = scenario, y = value, 
           group = interaction(groom_strategy, theta), 
           color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(groom_strategy~metric, ncol = 4, nrow=6, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()
pop_met_rg_ni_plot# check all good 

png(filename="figures/FigureX_rg_ni_prev.png", width = 16, height =8, unit = 'in', res = 300)
pop_met_rg_ni_plot 
dev.off()

# plot individual metrics for only resource-induced changes to grooming, and baseline immigration (0.2)

pop_met_rg_bi_plot <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0.2, ], 
       aes(x = scenario, y = value, 
           group = interaction(groom_strategy, theta), 
           color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(metric ~groom_strategy, ncol = 4, nrow=6, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()

pop_met_rg_bi_plot # check all good

# # plot individual metrics for only resource-induced changes to grooming, and baseline immigration (0.2) and mid theta (2) 

pop_met_rg_bimt_plot <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0.2 &
                                                       pop_stat_per_param_long$theta == 5, ], 
       aes(x = scenario, y = value, 
           color = groom_strategy)) +
  geom_point(size = 2) +
  facet_wrap(metric~groom_strategy, ncol = 4, nrow=6, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()

pop_met_rg_bimt_plot # check all good 

# save plot 
png(filename="figures/FigureX_rg_bimt_pop_metrics.png", width = 16, height =8, unit = 'in', res = 300)
pop_met_rg_bimt_plot 
dev.off()


## Plot separate panels for individual metrics for only resource-induced changes to grooming and no immigration
prev_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 & 
                                               pop_stat_per_param_long$metric == "pop_metric_prev",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta),  
                          color = groom_strategy)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4,1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "prevalence") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
prev_plot_ni

# remove legend for multi-panel plot: 
prev_plot_ni_nl <- prev_plot_ni +
  theme(legend.position = "none",
        axis.title.x = element_blank()) # use joint label instead

## average eod burden 
abund_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 &
                                                  pop_stat_per_param_long$metric ==
                                                  "pop_metric_avg_abund",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "abundance") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
abund_plot_ni

# remove legend for multi-panel plot: 
abund_plot_ni_nl <- abund_plot_ni +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) # use joint label instead

## Intensity  
intense_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 & 
                                                    pop_stat_per_param_long$metric == 
                                                            "pop_metric_inte",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta),
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4,1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total groomed population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
intense_plot_ni

# remove legend for multi-panel plot: 
intense_plot_ni_nl <- intense_plot_ni +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none")
  
## k aggregation
k_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 &
                                            pop_stat_per_param_long$metric == "pop_metric_k",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "aggegation (k)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank()) # use joint label instead
k_plot_ni # check all good 

# Multipanel plot for no immigration 
png(filename="figures/FigureX6_pop_metrics_ni.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(prev_ni_nl, abund_plot_ni_nl , intense_plot_ni_nl, k_plot_ni, 
             nrow = 1,
             bottom = "resource availability")
dev.off()

# Now Resources jointly impacts immigration and grooming  
## use resource pairs for just grooming 
  
  
pop_stat_per_param_long_ri_link <- pop_stat_per_param_long %>%
  inner_join(resource_pairs, by = c("scenario", "iota"))

# confirm grooming strategy levels
pop_stat_per_param_long_ri_link$groom_strategy <- factor(pop_stat_per_param_long_ri_link$groom_strategy,
                                                          levels = c("none", 
                                                                     "programmed", 
                                                                     "stimulus", 
                                                                     "costly")) 
# prepare individual metric plots 

# PLOT OF RESOURCE on individual metrics with immigration-grooming paired and Theta = 5
## total population 
prev_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                        pop_stat_per_param_long_ri_link$metric == "pop_metric_prev",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "prevalence") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
prev_plot

# remove legend for multi-panel plot: 
prev_plot_nl <- prev_plot +
  theme(legend.position = "none",
        axis.title.x = element_blank()) # use joint label instead

## average eod burden 
abund_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                                    pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_abun",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean abundance") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
abund_plot

# remove legend for multi-panel plot: 
abund_plot_nl <- abund_plot +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) # use joint label instead

## INTENSITY  
intense_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                                       pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_inte",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean intensity") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
intense_plot

# remove legend for multi-panel plot: 
intense_plot_nl <- intense_plot +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none")
  
## Aggregation k 
k_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                                 pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_k",], 
                      aes(x = scenario, y = round(value/780*100,2), 
                          group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  
  facet_wrap(~groom_strategy, ncol = 4) +
  geom_line(y= 1, color = "black") +
  geom_line(y= 0, color = "grey") +
  geom_line(y= 20, color = "grey") +
  scale_y_continuous(limits = c(0,20)) +
  labs(x = "resource availability",
       y = "aggregation (k)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank()) # use joint label instead
k_plot # check all good 

# Multipanel plot for one theta value (5)
png(filename="figures/FigureX_pop_metrics_rgi_t7.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(prev_plot_nl, abund_plot_nl , intense_plot_nl, k_plot, nrow = 4,
             bottom = "resource availability")
dev.off()

# PLOT OF RESOURCE on individual metrics with immigration-grooming paired with variable Theta
## total population for all thetas
prev_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_prev",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "prevalence") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
prev_plot_at 

# remove legend for multi-panel plot: 
prev_plot_at_nl <- prev_plot_at +
  theme(legend.position = "none",
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## average eod burden for all thetas
abund_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_abun",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean abundance") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
abund_plot_at

# remove legend for multi-panel plot: 
abund_plot_at_nl <- abund_plot_at +
  theme(legend.position = "none", 
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## Intensity  
intense_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_inte",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy,theta),
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean intensity") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
intense_plot_at

# remove legend for multi-panel plot: 
intense_plot_at_nl <- intense_plot_at +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none",
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
  
## k aggregation 
k_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_k",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy,theta),  
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  scale_y_continuous(limits = c(0,5)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "aggregation (k)") +
  geom_line(y= 1, color = "black") +
  geom_line(y= 0, color = "grey") +
  geom_line(y= 20, color = "grey") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank(),# use joint label instead
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
k_plot_at  # check all good 

# Save Plot 
png(filename="figures/FigureX7_pop_metrics_ri-link.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(prev_plot_at_nl, abund_plot_at_nl, intense_plot_at_nl, k_plot_at, nrow = 4,
             bottom = grid::textGrob("resource availability",
                               gp = grid::gpar(fontsize = 20, fontface = "bold")))
dev.off()

# Lets focus on just costly and stimulus, as they appear to only have suitable k variation, limit theta 
k_plot_at_sc_3t <- ggplot(pop_stat_per_param_long_ri_link[
  pop_stat_per_param_long_ri_link$metric == "pop_metric_k" &
  pop_stat_per_param_long_ri_link$groom_strategy %in% c("stimulus", "programmed") &
  pop_stat_per_param_long_ri_link$theta == 7,], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy,theta),  
                          color = groom_strategy, alpha = theta)) +
  geom_point(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  scale_y_continuous(limits = c(0,5)) +
  facet_wrap(~groom_strategy, ncol = 2) +
  labs(x = "resource availability",
       y = "aggregation (k)") +
  geom_line(y= 1, color = "black") +
  geom_line(y= 0, color = "grey") +
  geom_line(y= 20, color = "grey") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank(),# use joint label instead
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
k_plot_at_sc_3t 
```

```{r k heatmap}
library(scales)

k_hm_plot_df <- pop_stat_per_param_long %>%
  filter(
    scenario == 50,
    metric == "pop_metric_k"
  ) %>%
  mutate(
    fill_value = case_when(
      is.na(value) ~ -2,   # extinction
      value < 0    ~ -1,   # underdispersed
      value > 10   ~ 11,   # random
      TRUE         ~ value # 0–10 linear
    ),
    class = case_when(
      is.na(value) ~ "Extinct",
      value < 0    ~ "Underdispersed",
      value > 10   ~ "Random",
      TRUE         ~ "Aggregated"
    )
  )

k_gs_pct <- k_hm_plot_df %>%
  filter(class %in% c("Extinct", "Underdispersed", "Aggregated", "Random")) %>%
  count(groom_strategy, class) %>%
  group_by(groom_strategy) %>%
  mutate(prop = n / sum(n),
         class = factor(
      class,
      levels = c("Extinct", "Underdispersed", "Aggregated", "Random")
    )
  )

k_gs_pct_plot <- ggplot(k_gs_pct, aes(x = 1, y = prop, fill = class)) +
  geom_col(width = 0.8) +
  coord_flip() +
  facet_grid(~ groom_strategy) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.5, 1),
    labels = c("0", "0.5", "1")
  ) +
  scale_fill_manual(
    values = c(
      "Extinct" = "yellow",
      "Underdispersed" = "blue",
      "Aggregated" = "gray50",
      "Random" = "red"
    ),
    drop = FALSE,
    name = "Dispersion"
  ) +
  labs(x = NULL, y = "Proportion") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_blank(),
    legend.position = "none"
  )
k_gs_pct_plot # check all good

# Save proportion of k categories
png(filename="figures/FigureX_kpop_cats.png", width = 8, height =8, unit = 'in', res = 300)
k_gs_pct_plot
dev.off()

k_legend_df <- data.frame(
  class = c("Extinct", "Underdispersed", "Aggregated", "Random"),
  x = NA_real_,
  y = NA_real_
)

k_it <- ggplot(k_hm_plot_df, aes(x = theta, y = iota)) +
  geom_tile(aes(fill = fill_value)) +

  scale_fill_gradientn(
    colours = c(
      "yellow",  # extinction
      "blue",    # <0
      "black",   # 0
      "gray85",   # 10
      "red"      # >10
    ),
    values = rescale(c(
      -2,   # extinct
      -1,   # underdispersed
      0,    # start of gradient
      10,   # end of gradient
      11    # random
    )),
    limits = c(-2, 11),
    breaks = c(0, 2, 5, 10),
    labels = c("0", "2", "5", "10"),
    name = "k (0–10)"
  ) +

  # dummy legend for class
  geom_point(
    data = k_legend_df,
    aes(x = x, y = y, color = class),
    inherit.aes = FALSE,
    alpha = 0
  ) +

  scale_color_manual(
    values = c(
      "Extinct" = "yellow",
      "Underdispersed" = "blue",
      "Aggregated" = "gray50",
      "Random" = "red"
    ),
    name = "Class"
  ) +

  facet_grid(~ groom_strategy) +
  labs(x = "theta", y = "iota") +
  theme_minimal() +
  guides(
    color = guide_legend(order = 1, override.aes = list(alpha = 1, size = 5)),
    fill  = guide_colorbar(order = 2)
  )
k_it # confirm all good


# Save Heatmap
png(filename="figures/FigureX_kpop_sensitivity.png", width = 8, height =8, unit = 'in', res = 300)
k_it
dev.off()

# Filter to Aggregated scenarios 
agg_pop_stat_per_param <- pop_stat_per_param[pop_stat_per_param$scenario == 50 &
                               pop_stat_per_param$pop_metric_k < 1 & 
                               pop_stat_per_param$pop_metric_k > 0, ]


# Based on this, we will explore theta = 6 and iota = 0.05 as aggregated 
prev_plot_agg <- ggplot(pop_stat_per_param_long_ri_link[
  pop_stat_per_param_long_ri_link$metric == "pop_metric_prev" & 
  pop_stat_per_param_long_ri_link$theta == 6,], 
                      aes(x = scenario, y = value, group = groom_strategy, 
                          color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "prevalence") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
prev_plot_agg 

# remove legend for multi-panel plot: 
prev_plot_agg_nl <- prev_plot_agg +
  theme(legend.position = "none",
        axis.title.y = element_text(face = "bold", size =14),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## average eod burden for all thetas
abund_plot_agg <- ggplot(pop_stat_per_param_long_ri_link[
  pop_stat_per_param_long_ri_link$metric == "pop_metric_abun" &
  pop_stat_per_param_long_ri_link$theta == 6,], 
                      aes(x = scenario, y = value, group = groom_strategy, 
                          color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean abundance") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
abund_plot_agg

# remove legend for multi-panel plot: 
abund_plot_agg_nl <- abund_plot_agg +
  theme(legend.position = "none", 
        axis.title.y = element_text(face = "bold", size =14),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## Intensity  
intense_plot_agg <- ggplot(pop_stat_per_param_long_ri_link[
  pop_stat_per_param_long_ri_link$metric == "pop_metric_inte" &
  pop_stat_per_param_long_ri_link$theta == 6,], 
                      aes(x = scenario, y = value, group = groom_strategy,
                          color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean intensity") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
intense_plot_agg

# remove legend for multi-panel plot: 
intense_plot_agg_nl <- intense_plot_agg +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none",
        axis.title.y = element_text(face = "bold", size =14),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
  
## k aggregation 
k_plot_agg <- ggplot(pop_stat_per_param_long_ri_link[
  pop_stat_per_param_long_ri_link$metric == "pop_metric_k" & 
  pop_stat_per_param_long_ri_link$theta == 6,], 
                      aes(x = scenario, y = value, 
                          group = groom_strategy,  
                          color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "aggregation (k)") +
  geom_line(y= 10, color = "grey") +
  geom_line(y= 1, color = "black") +
  geom_line(y= 0, color = "grey") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank(),# use joint label instead
        axis.title.y = element_text(face = "bold", size =14),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
k_plot_agg  # check all good 

# Save Plot 
png(filename="figures/FigureX_agg_metrics_ri-link.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(prev_plot_agg_nl, abund_plot_agg_nl, intense_plot_agg_nl, k_plot_agg, nrow = 4,
             bottom = grid::textGrob("resource availability",
                               gp = grid::gpar(fontsize = 20, fontface = "bold")))
dev.off()

```
```{r}
# -- Plot 2: Strategy means + IQR ribbons --
summary_df <- results |>
  group_by(strategy, day) |>
  summarise(mean_N = mean(N),
            lo     = quantile(N, 0.25),
            hi     = quantile(N, 0.75),
            .groups = "drop")
```

