---
title: "Willoughby ectoparasite model"
author: "Anna Willoughby and John Drake"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=FALSE)
library('deSolve') # solving ordinary differential equations
library(dplyr) # data wrangling
library(ggplot2) # pretty plots
set.seed(1993)
```

## Model

In this document we seek to use a birth-death-immigration process to model the number of ectoparasites on a host.A full description of the model is at https://github.com/DrakeLab/willoughby-grooming-model.

We start by writing functions for the birth rate, death rate, and immigration rate.
In all cases, we let the state variable ($x$) be the first parameter.
We also set some default parameters to make things simpler later.


```{r, functions}
birthrate <- function(x, b0, theta) b0*x*(x/(theta+x)) # function for birth rate with Allee effect (theta)
deathrate <- function(x, d0, d1) (d0+d1*x)*x # function for death rate (quadratic right now?)
immigrationrate <- function (x, iota) iota # function for immigration rate
# set initial conditions parameters: birth rate, allee effect, small death rate, death rate at high pops 
# birth rate is standard across all scenarios, right now as 100 flea births per year (0.27 per day)
# set for each: 
## Basic Model (0)
parms0 <-  c(b0=0.27, theta=0, d0=0, d1=0.01, iota=0, initial_pop = 2) # no allee, no immigration
## Allee effect model (1)
parms1 <-  c(b0=0.27, theta=6, d0=0, d1=0.01, iota=0, initial_pop = 2) # no immigration
## Resources (2)
parms2 <-  c(b0=0.27, theta=0, d0=0, d1=0.01, iota=1, initial_pop = 2) # no allee
## Multiple Effects Model (3)
parms3 <-  c(b0=0.27, theta=6, d0=0, d1=0.01, iota=1, initial_pop = 2) 

# make a list of the different model scenarios
model_scenarios <- list(parms0, parms1, parms2, parms3)

```

Now, we study the deterministic model numerically, as this will help to choose parameter values.

```{r, deterministic model}
# define the function for population growth rate
f <- function(t, x, parms){
  dx <- birthrate(x, parms[1], parms[2]) - 
    deathrate(x, parms[3], parms[4]) + 
    immigrationrate (x, parms[5])
  
  return(list(dx))
}

# set initial conditions: start with one individual and run through all model scenarios 
## Model 0 
solution0 <- as.data.frame(ode(y=1, times=seq(1:365), func=f, parms=parms0))  
solution0$model <- "zero"
solution0$Allee = "no"
solution0$immigration = "no"
## Model 1
solution1 <- as.data.frame(ode(y=1, times=seq(1:365), func=f, parms=parms1)) 
solution1$model <- "one"
solution1$Allee = "yes"
solution1$immigration = "no"
## Model 2
solution2 <- as.data.frame(ode(y=1, times=seq(1:365), func=f, parms=parms2))  
solution2$model <- "two"
solution2$Allee = "no"
solution2$immigration = "yes"
## Model 3
solution3 <-  as.data.frame(ode(y=1, times=seq(1:365), func=f, parms=parms3))  
solution3$model <- "three"
solution3$Allee = "yes"
solution3$immigration = "yes"

# Combine all solutions into one dataframe
solutions <- bind_rows(solution0, solution1, solution2, solution3)

# Rename columns for clarity
colnames(solutions) <- c("time", "population", "model", "Allee", "immigration")

# Plot using ggplot
ggplot(solutions, aes(x = time, y = population, linetype = Allee, color = immigration, group = model)) +
  geom_line(size = 1) +  # Add lines
  theme_minimal() +  # Clean theme
  labs(title = "Deterministic Model Trajectories",
       x = "Time (days)",
       y = "Population Size",
       linetype = "Allee effect",
       color = "immigration") +
  scale_linetype_manual(values = c("solid", "dashed")) +  # Different line types
  scale_color_manual(values = c("black", "blue"))  # Custom colors

```

### solve for the carrying capacity, and max pop

```{r max pop calculation}
# when does f = 0 ? 

# Define the equation to solve # [can prop simplify with f above....]
equilibrium_eq <- function(x, parms) {
  with(as.list(parms), {
    birthrate <- b0*x*(x/(theta+x))
    deathrate <- (d0 + d1*x) * x
    immigrationrate <- iota
    return(birthrate - deathrate + immigrationrate)
  })
}

# Function to find equilibrium population
max_pop_solve <- function(parms) {
  root <- tryCatch(
    uniroot(function(x) equilibrium_eq(x, parms), c(0, 1000), tol = 1e-6)$root,
    error = function(e) NA  # Return NA if uniroot fails
  )
  return(root)
}
# Solve for maximum population for each scenario
max_pops <- lapply(model_scenarios, max_pop_solve)

# Print results
names(max_pops) <- c("Basic Model", "Allee Effect Model", "Resources Model", "Multiple Effects Model")
print(max_pops)
```
## Stochastic Model

Now, we want to simulate a birth-death-immigration process.
We start with some hyperparameters of the simulation.

```{r set some simulation rules}
num_simulations <- 20  # Number of times to run the simulation
# max.events <- 1000 # maximum events
max.time <- 730 # 2 year max for a squirrel  
```

Here we simulate the process


```{r run stochastic models}
num_simulations <- 20  # Define number of simulations
max.time <- 730  # Define max time

# 0: Run basic model 
simulation0 <- vector("list", num_simulations)  # Initialize storage

  for (sim in 1:num_simulations) {
    events <- 0  # Initialize the total number of events
    time <- 0  # Initialize current time
    x <- parms0["initial_pop"]  # Start population at initial value

    # Dataframe to store results
    simulation_results <- data.frame(event = numeric(),
                                     time = numeric(),
                                     population = numeric(), 
                                     event_type = character(),
                                     sim_id = numeric())

    while(time[length(time)] <= max.time) {  
      # Track number of events
      events <- events + 1  
      current.size <- x[length(x)]  # Use last recorded population size
      current.time <- time[length(time)]  # Use last recorded time
      
      # Calculate event rates with proper parameters
      b <- birthrate(x=current.size, b0=parms0["b0"], theta=parms0["theta"])
      d <- deathrate(x=current.size, d0=parms0["d0"], d1=parms0["d1"])
      i <- immigrationrate(x=current.size, iota=parms0["iota"])
      
      # Compute total rate and time to next event
      total.rate <- b + d + i
      if (total.rate == 0 || is.na(total.rate)) {
        break  # Stop simulation if no events can occur
        }
      increment.time <- -log(1 - runif(1)) / total.rate  # Exponential waiting time
      
      # Determine event type
      event_probs <- c(d, b, i) / total.rate
      event_outcome <- sample(c("death", "birth", "immigration"), size=1, prob=event_probs)
      change <- ifelse(event_outcome == "death", -1, ifelse(event_outcome == "birth", 1, 0))  # Immigration adds 0
      
      # Store results
      simulation_results <- rbind(simulation_results, data.frame(
        event = events,
        time = current.time + increment.time,
        population = max(0, current.size + change),  # Ensure population doesn't go negative
        event_type = event_outcome,
        sim_id = sim
      ))

      # Update state variables
      time <- c(time, current.time + increment.time)
      x <- c(x, max(0, current.size + change))  # Prevent negative population
    }

    # Store results of this simulation
    simulation0[[sim]] <- simulation_results
  }

# Save all results
saveRDS(simulation0, "data/simulation0.Rdata")

# 1: run Allee effect model 
simulation1 <- vector("list", num_simulations)  # Initialize storage

  for (sim in 1:num_simulations) {
    events <- 0  # Initialize the total number of events
    time <- 0  # Initialize current time
    x <- parms1["initial_pop"]  # Start population at initial value

    # Dataframe to store results
    simulation_results <- data.frame(event = numeric(),
                                     time = numeric(),
                                     population = numeric(), 
                                     event_type = character(),
                                     sim_id = numeric())

    while(time[length(time)] <= max.time) {  
      # Track number of events
      events <- events + 1  
      current.size <- x[length(x)]  # Use last recorded population size
      current.time <- time[length(time)]  # Use last recorded time
      
      # Calculate event rates with proper parameters
      b <- birthrate(x=current.size, b0=parms1["b0"], theta=parms1["theta"])
      d <- deathrate(x=current.size, d0=parms1["d0"], d1=parms1["d1"])
      i <- immigrationrate(x=current.size, iota=parms1["iota"])
      
      # Compute total rate and time to next event
      total.rate <- b + d + i
      if (total.rate == 0 || is.na(total.rate)) {
        break  # Stop simulation if no events can occur
        }
      increment.time <- -log(1 - runif(1)) / total.rate  # Exponential waiting time
      
      # Determine event type
      event_probs <- c(d, b, i) / total.rate
      event_outcome <- sample(c("death", "birth", "immigration"), size=1, prob=event_probs)
      change <- ifelse(event_outcome == "death", -1, ifelse(event_outcome == "birth", 1, 0))  # Immigration adds 0
      
      # Store results
      simulation_results <- rbind(simulation_results, data.frame(
        event = events,
        time = current.time + increment.time,
        population = max(0, current.size + change),  # Ensure population doesn't go negative
        event_type = event_outcome,
        sim_id = sim
      ))

      # Update state variables
      time <- c(time, current.time + increment.time)
      x <- c(x, max(0, current.size + change))  # Prevent negative population
    }

    # Store results of this simulation
    simulation1[[sim]] <- simulation_results
  }

  # Save all results
  saveRDS(simulation1, "data/simulation1.Rdata")

# 2: Run resources model
simulation2 <- vector("list", num_simulations)  # Initialize storage

  for (sim in 1:num_simulations) {
    events <- 0  # Initialize the total number of events
    time <- 0  # Initialize current time
    x <- parms2["initial_pop"]  # Start population at initial value

    # Dataframe to store results
    simulation_results <- data.frame(event = numeric(),
                                     time = numeric(),
                                     population = numeric(), 
                                     event_type = character(),
                                     sim_id = numeric())

    while(time[length(time)] <= max.time) {  
      # Track number of events
      events <- events + 1  
      current.size <- x[length(x)]  # Use last recorded population size
      current.time <- time[length(time)]  # Use last recorded time
      
      # Calculate event rates with proper parameters
      b <- birthrate(x=current.size, b0=parms2["b0"], theta=parms2["theta"])
      d <- deathrate(x=current.size, d0=parms2["d0"], d1=parms2["d1"])
      i <- immigrationrate(x=current.size, iota=parms2["iota"])
      
      # Compute total rate and time to next event
      total.rate <- b + d + i
      if (total.rate == 0 || is.na(total.rate)) {
        break  # Stop simulation if no events can occur
        }
      increment.time <- -log(1 - runif(1)) / total.rate  # Exponential waiting time
      
      # Determine event type
      event_probs <- c(d, b, i) / total.rate
      event_outcome <- sample(c("death", "birth", "immigration"), size=1, prob=event_probs)
      change <- ifelse(event_outcome == "death", -1, ifelse(event_outcome == "birth", 1, 0))  # Immigration adds 0
      
      # Store results
      simulation_results <- rbind(simulation_results, data.frame(
        event = events,
        time = current.time + increment.time,
        population = max(0, current.size + change),  # Ensure population doesn't go negative
        event_type = event_outcome,
        sim_id = sim
      ))

      # Update state variables
      time <- c(time, current.time + increment.time)
      x <- c(x, max(0, current.size + change))  # Prevent negative population
    }

    # Store results of this simulation
    simulation2[[sim]] <- simulation_results
  }

  # Save all results
  saveRDS(simulation2, "data/simulation2.Rdata")

# 3: Run complex model 
simulation3 <- vector("list", num_simulations)  # Initialize storage

  for (sim in 1:num_simulations) {
    events <- 0  # Initialize the total number of events
    time <- 0  # Initialize current time
    x <- parms3["initial_pop"]  # Start population at initial value

    # Dataframe to store results
    simulation_results <- data.frame(event = numeric(),
                                     time = numeric(),
                                     population = numeric(), 
                                     event_type = character(),
                                     sim_id = numeric())

    while(time[length(time)] <= max.time) {  
      # Track number of events
      events <- events + 1  
      current.size <- x[length(x)]  # Use last recorded population size
      current.time <- time[length(time)]  # Use last recorded time
      
      # Calculate event rates with proper parameters
      b <- birthrate(x=current.size, b0=parms3["b0"], theta=parms3["theta"])
      d <- deathrate(x=current.size, d0=parms3["d0"], d1=parms3["d1"])
      i <- immigrationrate(x=current.size, iota=parms3["iota"])
      
      # Compute total rate and time to next event
      total.rate <- b + d + i
      if (total.rate == 0 || is.na(total.rate)) {
        break  # Stop simulation if no events can occur
        }
      increment.time <- -log(1 - runif(1)) / total.rate  # Exponential waiting time
      
      # Determine event type
      event_probs <- c(d, b, i) / total.rate
      event_outcome <- sample(c("death", "birth", "immigration"), size=1, prob=event_probs)
      change <- ifelse(event_outcome == "death", -1, ifelse(event_outcome == "birth", 1, 0))  # Immigration adds 0
      
      # Store results
      simulation_results <- rbind(simulation_results, data.frame(
        event = events,
        time = current.time + increment.time,
        population = max(0, current.size + change),  # Ensure population doesn't go negative
        event_type = event_outcome,
        sim_id = sim
      ))

      # Update state variables
      time <- c(time, current.time + increment.time)
      x <- c(x, max(0, current.size + change))  # Prevent negative population
    }

    # Store results of this simulation
    simulation3[[sim]] <- simulation_results
  }

  # Save all results
  saveRDS(simulation3, "data/simulation3.Rdata")

####  List of all simulation models
all_simulations <- list(simulation0, simulation1, simulation2, simulation3)
names(all_simulations) <- c("Model 0", "Model 1", "Model 2", "Model 3")
```


```{r set up simulation}
# Store results in a list
simulations <- vector("list", num_simulations)

for (sim in 1:num_simulations) {

events <- 0 #initialize the total number of events
time <- 0  #initialize the current time
x <- 0 #initialize population size

# Create a dataframe to store results
simulation_results <- data.frame(event = numeric(),
                                 time = numeric(),
                                 population = numeric(), 
                                 event_type = character(),
                                 sim_id = numeric())
# no longer doing event length (events <= max.events)
while(tail(time,1) <= max.time ){
  
  # some bookkeeping
  events <- events + 1 #increment the number of events
  current.size <- x[events] #set local current population size
  current.time <- time[events]
  
  # get the rates and store locally
  b <- birthrate(x=current.size)
  d <- deathrate(x=current.size)
  i <- immigrationrate(x=current.size)
  
  # find the next event time
  total.rate <- b+d+i
  
  increment.time <- -log(1-runif(1))/total.rate # exponential time to next event
  
  # find next event type John's way
  # change <- sample(c(-1, 1), size=1, prob = c(d/total.rate,(b+i)/total.rate))
  
  # Determine event type Anna's way
  event_probs <- c(d, b, i) / total.rate # what is the rate right now for each of the event types? 
  event_outcome <- sample(c("death", "birth", "immigration"), size = 1, prob = event_probs) # sample what happens 
  change <- ifelse(event_outcome == "death", -1, 1) # what will be the population change? 
  
  # Store results in dataframe
  simulation_results <- rbind(simulation_results, data.frame(event = events,
                                                             time = current.time + increment.time,
                                                             population = current.size + change,
                                                             event_type = event_outcome,
                                                             sim_id = sim))
  # update state variables and store
  time <- c(time, current.time+increment.time)
  x <- c(x,current.size+change)
}

# Store this simulation's results
  all_simulations[[sim]] <- simulation_results
}
all_simulations <- saveRDS(all_simulations, "data/all_simulations.Rdata")
}

# Combine all simulation results into one dataframe
# final_results <- do.call(rbind, all_simulations)

# plot(time,x, type = 'b', xlab='Time', ylab='Ectoparasite population size') # this is for single sim 

library(ggplot2)
# make a plot with all the simulations 
# ggplot(final_results, aes(x = time, y = population, group = sim_id, color = factor(sim_id))) +
#  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
#  labs(title = "Ectoparasite Population Over Time (*n* = 100 simulations)", 
#       x = "Time", y = "Population Size", color = "Simulation ID") +
#  theme_classic() +
#  theme(legend.position = "none")  # hide legend cause too many lines
```
Now, I need a way to sample from these data.

```{r plot stochastic model output}

# Combine simulation results into one dataframe
results0 <- do.call(rbind, simulation0)
results1 <- do.call(rbind, simulation1)
results2 <- do.call(rbind, simulation2)
results3 <- do.call(rbind, simulation3)

# plot(time,x, type = 'b', xlab='Time', ylab='Ectoparasite population size') # this is for single sim 

library(ggplot2)
# make a plot with all the simulations 
ggplot(results0, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(title = "Ectoparasite Population Over Time (*n* = 20 simulations)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none")  # hide legend cause too many lines
ggplot(results1, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(title = "Ectoparasite Population Over Time (*n* = 20 simulations)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none")  # hide legend cause too many lines
ggplot(results2, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(title = "Ectoparasite Population Over Time (*n* = 20 simulations)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none")  # hide legend cause too many lines
ggplot(results3, aes(x = time, y = population, group = sim_id)) +
  geom_line(alpha = 0.3) +  # Add transparency to avoid clutter
  labs(title = "Ectoparasite Population Over Time (*n* = 20 simulations)", 
       x = "Time", y = "Population Size", color = "Simulation ID") +
  theme_classic() +
  theme(legend.position = "none")  # hide legend cause too many lines

```
```{r calculate traits - total pop}
# FLEA POPULATION TRAITS - calculate relevant traits 

## values for one simulation
### Total Flea Population: the count of fleas during the entire simulation run 

#### Function to calculate total flea population from immigration & birth events
calculate_flea_population <- function(simulation_list) {
  sapply(simulation_list, function(sim_df) {
    sum(sim_df$event_type == "immigration") + sum(sim_df$event_type == "birth")
  })
}

#### Apply function to all models
flea_total_pop_lists <- lapply(all_simulations, calculate_flea_population)

#### Plot histograms for each model as facet
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid
for (i in seq_along(flea_total_pop_lists)) {
  hist(flea_total_pop_lists[[i]], 
       main = paste("Total Flea Pop -", names(flea_total_pop_lists)[i]), 
       xlab = "Total Fleas", 
       col = "lightblue", 
       border = "black")
}
par(mfrow = c(1,1))  # Reset plot layout

#### Convert list to long-format dataframe for ggplot
flea_data <- bind_rows(
  lapply(names(flea_total_pop_lists), function(model) {
    data.frame(Total_Fleas = flea_total_pop_lists[[model]], Model = model)
  })
)

#### Plot histogram using ggplot2 for each model as overlapping on same grid
ggplot(flea_data, aes(x = Total_Fleas, fill = Model)) +
  geom_histogram(aes(y = ..density..), position = "identity", alpha = 0.4, bins = 30) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9","grey", "darkmagenta")) +  # Custom colors
  labs(title = "Total Flea Population Across Simulations",
       x = "Total Fleas from Immigration & Birth",
       y = "Density") +
  theme_minimal()

```

```{r calculate traits - snapshot pops}
### Snapshot/Sampling Event Flea Population: the count of fleas at one snapshot (e.g., end of day)

#### create function round time to the nearest integer, keep only the last observation per time unit
end_of_day_flea_pop <-function(sim){
  sim %>%  # Apply operations to the dataset
  mutate(time_unit = floor(time)) %>%  # Convert time to integer units
  group_by(time_unit) %>%
  slice_tail(n = 1) %>%  # Get the last observation per unit
  ungroup() %>%
  select(time_unit, population)  # Keep relevant columns
}

# apply the end of day calculation across simulations 
sim0_end_of_day_pops <- lapply(simulation0, end_of_day_flea_pop)
sim1_end_of_day_pops <- lapply(simulation1, end_of_day_flea_pop)
sim2_end_of_day_pops <- lapply(simulation2, end_of_day_flea_pop)
sim3_end_of_day_pops <- lapply(simulation3, end_of_day_flea_pop)

# create a function to sample only a portion of the days (e.g. last 25%)
subset_simulation <- function(sim_df, range_percent) {
  # Ensure range_percent is a valid numeric vector of length 2 (e.g., c(0, 10) or c(75, 100))
  if (length(range_percent) != 2 || any(range_percent < 0) || any(range_percent > 100)) {
    stop("range_percent must be a numeric vector of length 2 between 0 and 100.")
  }
  
  # Get the start and end time from the simulation
  min_time <- min(sim_df$time_unit)
  max_time <- max(sim_df$time_unit)

  # Compute the time range corresponding to the percentage range
  time_start <- min_time + (range_percent[1] / 100) * (max_time - min_time)
  time_end <- min_time + (range_percent[2] / 100) * (max_time - min_time)

  # Subset the dataframe based on the calculated time range
  sim_df %>%
    filter(time_unit >= time_start & time_unit <= time_end)
}

# limit population snapshots to latter 75% (eliminating burn-in period)
sim0_eod_pop_last75 <- lapply(sim0_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim0_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim0_eod_pop_last75), function(i) {
    sim0_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim0_eod_pop_last75_combined$model <- "0 - basic"
sim0_eod_pop_last75_combined$sim_id <- paste("0-", sim0_eod_pop_last75_combined$sim_id, sep = "")
sim1_eod_pop_last75 <- lapply(sim1_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim1_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim1_eod_pop_last75), function(i) {
    sim1_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim1_eod_pop_last75_combined$model <- "1 - allee"
sim1_eod_pop_last75_combined$sim_id <- paste("1-", sim1_eod_pop_last75_combined$sim_id, sep = "")

sim2_eod_pop_last75 <- lapply(sim2_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim2_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim2_eod_pop_last75), function(i) {
    sim2_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim2_eod_pop_last75_combined$model = "2 - resources"
sim2_eod_pop_last75_combined$sim_id <- paste("2-", sim2_eod_pop_last75_combined$sim_id, sep = "")

sim3_eod_pop_last75 <- lapply(sim3_end_of_day_pops, subset_simulation, range_percent = c(26, 100))
sim3_eod_pop_last75_combined <- bind_rows(
  lapply(seq_along(sim3_eod_pop_last75), function(i) {
    sim3_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
  })
)
sim3_eod_pop_last75_combined$model = "3 - complex"
sim3_eod_pop_last75_combined$sim_id <- paste("3-", sim3_eod_pop_last75_combined$sim_id, sep = "")

eod_pop_last75 <- rbind(sim0_eod_pop_last75_combined, 
                        sim1_eod_pop_last75_combined, 
                        sim2_eod_pop_last75_combined, 
                        sim3_eod_pop_last75_combined)
# plot the histogram 
# hist(sim0_eod_pop_last75[[1]]$population)
ggplot(eod_pop_last75, aes(x = population, group = sim_id, color = model)) +
 # geom_freqpoly(binwidth = 10, size = 1, alpha = 0.5) +  # stepbins like a histogram
  geom_histogram(alpha = 0.2, aes(y = ..ncount..), position = 'identity', bins = 10) + 
 # geom_bar(pos="dodge") +
  scale_x_continuous(limits = c(0, 50)) +
 # scale_y_continuous(limits = c(0, 0.15)) +
  labs(title = "Distribution of Population Sizes for last 75% of simulation (n = 541 time points) ",
       x = "Population Size",
       y = "Frequency") +
  theme_classic() + 
 facet_grid(model ~ .)
```

```{r calculate aggregation metric - variance to mean ratio }
source("functions/Lc.R")

Gini <- function(x, corr = FALSE, na.rm = TRUE)
{
    if(!na.rm && any(is.na(x))) return(NA_real_)
    x <- as.numeric(na.omit(x))
    n <- length(x)
    x <- sort(x)
    G <- sum(x * 1L:n)
    G <- 2 * G/sum(x) - (n + 1L)
    if (corr) G/(n - 1L) else G/n
}

calc_agg_metrics <- function(df){
  mu <- mean(df$population)  # Mean of population size
  var <- var(df$population)  # Variance of population size
  var2mu <- var / mu # dispersion
  k <-   mu^2 / (var - mu) # aggregation parameter from negative binomial 
  Gini <- Gini(df$population) # Gini coef calc from ineq package
  return(data.frame(mu = mu, var = var, var2mu = var2mu, k = k, Gini = Gini))
}
# test out on a single df 
# sim02 <- calc_agg_metrics(sim0_eod_pop_last75[[2]])

# Apply function to each simulation and combine results
## Model 0 
sim0_agg_metrics_results <- lapply(sim0_eod_pop_last75, calc_agg_metrics)
sim0_agg_metrics <- do.call(rbind, sim0_agg_metrics_results)
sim0_agg_metrics$sim_id <- seq.int(nrow(sim0_agg_metrics))
sim0_agg_metrics$sim_id <- paste("model0",sim0_agg_metrics$sim_id, sep = "-")
## Model 1
sim1_agg_metrics_results <- lapply(sim1_eod_pop_last75, calc_agg_metrics)
sim1_agg_metrics <- do.call(rbind, sim1_agg_metrics_results)
sim1_agg_metrics$sim_id <- seq.int(nrow(sim1_agg_metrics))
sim1_agg_metrics$sim_id <- paste("model1",sim1_agg_metrics$sim_id, sep = "-")
## Model 2 
sim2_agg_metrics_results <- lapply(sim2_eod_pop_last75, calc_agg_metrics)
sim2_agg_metrics <- do.call(rbind, sim2_agg_metrics_results)
sim2_agg_metrics$sim_id <- seq.int(nrow(sim2_agg_metrics))
sim2_agg_metrics$sim_id <- paste("model2",sim2_agg_metrics$sim_id, sep = "-")
## Model 3
sim3_agg_metrics_results <- lapply(sim3_eod_pop_last75, calc_agg_metrics)
sim3_agg_metrics <- do.call(rbind, sim3_agg_metrics_results)
sim3_agg_metrics$sim_id <- seq.int(nrow(sim3_agg_metrics))
sim3_agg_metrics$sim_id <- paste("model3",sim3_agg_metrics$sim_id, sep = "-")

agg_metrics <- rbind(sim0_agg_metrics, 
                     sim1_agg_metrics, 
                     sim2_agg_metrics, 
                     sim3_agg_metrics)
# View the combined data frame
print(agg_metrics_df)
# View results
print(agg_metrics_results)

# need to test if any flea pops go extinct

# crowding

## HOST TRAITS 
## need to fix time units --- currently in flea reproduction units..... 
# calculate mean daily end in flea pop -- > this is the intensity! 


# across a single simulation 
## calculate prevalence: # per day? how many days does the host have an infection? 

## calculate abundance
### how many total fleas across the simulation 
#### how a daily metric of fleas and thean a mean metric? 


# across all simulations 
## calculate prevalence: for how many simulations does the population go extinct? 


# mean abundance

# mean intensity
```



JOHNS STUFF BELOW 
```{r, sampling-function}
# FLEA POPULATION TRAITS 
## Flea poplation at one-time point 
# create a function to sample the simulation at specific time points 

sample_population <- function(event_times, population_sizes, sample_times, initial_pop = NA) {

    # Ensure event_times and population_sizes are vectors of the same length
  if (length(event_times) != length(population_sizes)) {
    stop("event_times and population_sizes must be vectors of the same length.")
  }
  
  # Ensure event_times are sorted in increasing order
  if (is.unsorted(event_times)) {
    stop("event_times must be sorted.")
  }
  
  # Find the interval indices for each sampling time
  indices <- findInterval(sample_times, event_times)
  
  # Return the value of x from the sample intervals
  result <- population_sizes[indices]

  return(result)
}
```

Here I demonstrate use of this function.

```{r, sample-histogram}
s <- sample_population(time, x, seq(0.1,8, by=0.01)) 
hist(s, main='Histogram of ectoparasite burden', xlab='Individuals')
```

Now, let's create a version of the model that runs until the age limit of a host.
First we have to figure out how long a host (rock squirrel) lives in units of flea reproduction.
We will suppose a rock squirrel lives for five years and that a flea reproduces 100 off spring in one year.
That means that a rock squirrel lives through 500 reproduction units of fleas.
If the *per capita* flea birth rate is 1.0, then on average a rock squirrel lives to time 500.
(Note: this could be slow to compute.)

```{r max age of host in flea reproduction times}
max.time <- 500 #later change to 500
max.events <- 1000
```

Here we simulate the process

```{r write the simulation}
events <- 0 #initialize the total number of events
time <- 0  #initialize the current time
x <- 0 #initialize population size

while(tail(time,1) <= max.time ){
  
  # some bookkeeping
  events <- events + 1 #increment the number of events
  current.size <- tail(x,1)  # Always take the last value
  current.time <- tail(time,1) # Always take the last value
  
  # get the rates and store locally
  b <- birthrate(x=current.size)
  d <- deathrate(x=current.size)
  i <- immigrationrate(x=current.size)
  
  # find the next event time
  total.rate <- b+d+i
  
  increment.time <- -log(1-runif(1))/total.rate # exponential time to next event
  
  # find next event type
  change <- sample(c(-1, 1), size=1, prob = c(d/total.rate,(b+i)/total.rate))
  
  # update states and store
  time <- c(time, current.time+increment.time)
  x <- c(x,current.size+change)
}

plot(time,x, type = 'b', xlab='Time', ylab='Ectoparasite population size')
```

Here we write the simulation loop in a function that allows us to simulate the trajectory of infection in an individual of age $A$.

We will also generalize the function to arbitrary input parameters so that we don't rely on the defaults.

```{r}
simulate <- function(A = 500,
                     parms = NULL,
                     plot = FALSE) {
  events <- 0 #initialize the total number of events
  time <- 0  #initialize the current time
  x <- 0 #initialize population size
  
  if (is.null(parms)) {
    b0 <- 1
    theta <- 1
    d0 <- 0
    d1 <- 0.01
    iota <- 1
    
  } else {
    
    b0 <- parms$b0
    theta <- parms$theta
    d0 <- parms$d0
    d1 <- parms$d1
    iota <- parms$iota
    
  }
  
  while (tail(time, 1) <= A) {
    # some bookkeeping
    events <- events + 1 #increment the number of events
    current.size <- x[events] #set local current population size
    current.time <- time[events]
    
    # get the rates and store locally
    b <- birthrate(x = current.size, b0=b0, theta=theta)
    d <- deathrate(x = current.size, d0=d0, d1=d1)
    i <- immigrationrate(x = current.size, iota=iota)
    
    # find the next event time
    total.rate <- b + d + i
    
    increment.time <-
      -log(1 - runif(1)) / total.rate # exponential time to next event
    
    # find next event type
    change <-
      sample(c(-1, 1),
             size = 1,
             prob = c(d / total.rate, (b + i) / total.rate))
    
    # update states and store
    time <- c(time, current.time + increment.time)
    x <- c(x, current.size + change)
  }
  
  if (plot == T) {
    plot(time,
         x,
         type = 'b',
         xlab = 'Time',
         ylab = 'Ectoparasite population size')
  }
  
  return(data.frame(time, x))
  
}
```
Here we apply the function to an individual that lives 100 time units (one year)

```{r}
out1 <- simulate(A=100, plot=TRUE)
```

Here we apply the function to a different set of parameters (much smaller immigration rate).

```{r}
parameters <- list(b0=1, theta=1, d0=0, d1=0.01, iota=0.02)
out2 <- simulate(A=100, parms = parameters, plot=TRUE)
```

