---
title: "Workflow-v2"
author: "Anna Willoughby"
date: "2025-05-22"
output: pdf_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library('deSolve') # solving ordinary differential equations
library(dplyr) # data wrangling
library(ggplot2) # pretty plots
library(lhs) # latin hypercube sampling
library(tidyr) # pipes
library(parallel) # for running in parallel 
library(gridExtra) # plotting multipanels 
set.seed(1993)

```

## Model

In this document we seek to use a birth-death-immigration process to model the number of ectoparasites on a host. A full description of the model is at https://github.com/DrakeLab/willoughby-grooming-model.

We start by writing functions for the birth rate, death rate, and immigration rate.
In all cases, we let the state variable ($x$) be the first parameter.
We also set some default parameters to make things simpler later.


```{r, functions}

# function for birth rate with Allee effect (theta)
birthrate <- function(x, b0, theta, K){
  b <- b0*x*(x/(theta+x)) # baseline births with Allee
  b_dd <- b*x*(1 - x/ K) # birth decreases at high densities
  b_wA_dd <- pmax(0, b_dd)
  b_wA_dd # birth rate with Allee and density dependence
}

immigrationrate <- function (x, iota) iota # function for immigration rate (iota)

# read in all the death rates 

## Define all Grooming Scenarios: 
### PROGRAMMED 
groom_prg_scenarios <- data.frame(scenario = seq(1, 100, length.out = 100), # 100 scenarios 
                                  G_vals = seq(1, 300, length.out = 100)) # grooming efficiency
#### Generate a function for grooming deathrate based on pop size 
groomrate_prg <- function(pop.size, scenario){
  groom_prg_scenario <- groom_prg_scenarios %>% dplyr::filter(scenario == scenario) # filter to scenario
  death_rate <- pmin(pop.size, groom_prg_scenario[1,2]) # how much grooming will happen
}

### STIMULUS
groom_stm_scenarios <- data.frame(scenario = seq(1, 100, length.out = 100), # 100 scenarios 
                                  G_vals = seq(1, 100, length.out = 100)) # grooming efficiency
groom_stm_scenarios$groom_rate <- groom_stm_scenarios$G_vals / 100
  
#### Generate a function for stimulus grooming deathrate based on pop size 
groomrate_stm <- function(pop.size, scenario){
  death_rate <- ifelse(pop.size == 0, 0, # avoids NaNs
                       pop.size*groom_stm_scenarios[scenario,3])
  death_rate
}

### Costly
n_scen <-  100
# max grooming of 80 per day 
groom_cst_scenarios <- data.frame(
  scenario = 1:n_scen,
  mu = seq(10, 80, length.out = n_scen),
  sigma = seq(2, 20, length.out = n_scen)
) # max var of 20 sigma

#### Generate a function for costly grooming deathrate based on pop size 
groomrate_cst <- function(pop.size, scenario){
  groom_cst_scenario <- groom_cst_scenarios %>% dplyr::filter(scenario == scenario) # filter to scenario 
  death_rate <- ifelse(pop.size == 0, 0, # avoids NaNs
                       pop.size*exp(-((pop.size - groom_cst_scenario$mu)^2) / 
                                      (2 * groom_cst_scenario$sigma^2))) # normal distribution
  death_rate
}

# Function to select a groom strategy based on defined parameter: 
get_groom_rate <- function(x, groom_strategy, scenario) {
  
 ifelse(groom_strategy == "none", 0,
 ifelse(groom_strategy == "programmed", groomrate_prg(x, scenario),
 ifelse(groom_strategy == "stimulus", groomrate_stm(x, scenario),
 ifelse(groom_strategy == "costly", groomrate_cst(x, scenario), "Unknown groom_type"))))
}

# Function for density dependent deaths 
deathrate_dd <- function(pop.size, dmax = 1, x_mid = 50, s = 20) {
     dmax / (1 + exp(-(pop.size - x_mid) / s))
}

# set initial conditions parameters

## birth rate is standard across all scenarios, right now as 37.5 flea births per year (0. per day)
## we are going to vary parameters for:  
theta_vals <- seq(0, 10, by = 1) # 10 different values
iota_vals <- seq(0, 0.5, by = 0.05) # 10 different values
scenario_vals <- seq(1,100, by = 1) # 100 different values (no grooming is an explicit scenario) 

# construct full matrix of theta and iota options 
ti_axes <- expand.grid(
  theta = theta_vals, 
  iota = iota_vals, 
  scenario = scenario_vals
)

# Expand parameter set to include each grooming scenario 
groom_strategies <- c("none", "programmed", "stimulus", "costly")
tig_axes <- merge(ti_axes,
                  data.frame(groom_strategy = groom_strategies),
                             by = NULL)


# n_samples <- 500  # number of parameter combos
# lhs_matrix <- randomLHS(n_samples, 3)  # for theta, iota

# Rescale each column to your parameter ranges
# theta_vals <- lhs_matrix[,1] * 10        #  allee effect (none to high (n = 10))
# iota_vals  <- lhs_matrix[,2] * 5        # immigration (none to high (n = 5))
# scenario_vals <- floor(lhs_matrix[,3] * 100)    # which grooming scenario 

# Construct parameter set
param_grid <- tig_axes
param_grid$b0 <- log(78.75)/89/2 # constant [log(female fecundity per lifetime)/generation time]
param_grid$initial_pop <- 1
```


## Stochastic Model

Now, we want to simulate a birth-death-immigration process.
We start with some hyperparameters of the simulation.

```{r set some simulation rules}
# Individual Experiment 
num_simulations <- 10  # Number of times to run the simulation
# max.events <- 1000 # maximum events
max.time <- 730 # 2 year max for a squirrel  
```

Here we simulate the process

```{r run stochastic models}
# g_scenario <- read.csv(file = "data/grooming_deathrates.csv")
# groom_prog <- g_scenario %>% dplyr::filter(groom_type == "programmed")

# Define the task grid: 
task_grid <- expand.grid(
    param_id = 1:nrow(param_grid),
    sim_id   = 1:num_simulations)

# Define running the whole model as a function 
run_one_parameter_set <- function(task_i){

    param_id <- task_grid[task_i, 1]
    sim_id   <- task_grid[task_i, 2]
    
    message("Parameter set ", param_id, ", Simulation", sim_id)
    
    parms <- param_grid[param_id, ]

    events <- 0
    time <- 0
    x <- parms$initial_pop

    out_list <- list()

    while (time[length(time)] <= max.time) {

      events <- events + 1
      current.size <- x[length(x)]
      current.time <- time[length(time)]

      b <- birthrate(current.size, parms$b0, parms$theta, 50) # 50 as K
      g <- get_groom_rate(current.size, parms$groom_strategy, parms$scenario)
      #message("Current size", current.size, ", groom rate", g) # mid function checks
      d <- deathrate_dd(current.size) 
      iota <- immigrationrate(current.size, iota = parms$iota)

      total.rate <- b + g + d + iota

      if (is.na(total.rate) || total.rate <= 0) break

      increment.time <- -log(1 - runif(1)) / total.rate

      event_probs <- c(d, g, b, iota) / total.rate
      event_outcome <- sample(
        c("death","groomed","birth", "immigration"),
        size = 1,
        prob = event_probs
      )

      change <- ifelse(event_outcome %in% c("death", "groomed"), -1, 1)

      new_size <- max(0, current.size + change)
      new_time <- current.time + increment.time

      out_list[[events]] <- data.frame(
        event = events,
        time = new_time,
        population = new_size,
        event_type = event_outcome,
        sim_id = sim_id,
        param_id = param_id
      )

      time <- c(time, new_time)
      x <- c(x, new_size)
    }

    if (length(out_list) == 0) {
      data.frame(
        event = numeric(),
        time = numeric(),
        population = numeric(),
        event_type = character(),
        sim_id = numeric(),
        param_id = numeric()
      )
    } else {
      do.call(rbind, out_list)
}}

# First test out on one [no longer needed]
# system.time({
#   simulation_test <-run_one_parameter_set(2565)
# }
# )

# ggplot(simulation_test, aes(x = time, y = population)) + 
#         geom_line()
  
# Now run for the full simulation across the 10 clusters 
system.time({

  cl <- makeCluster(10)
  
  # Move libraries to the clusters 
  clusterEvalQ(cl, library(tidyverse))

  # Move main environment items to each of the clusters
  clusterExport(cl, 
                c("groom_prg_scenarios","groom_stm_scenarios", "groom_cst_scenarios",
                   "param_grid","max.time", "task_grid", 
                   "birthrate", "deathrate_dd", "immigrationrate", 
                  "groom_strategies", "get_groom_rate", "groomrate_prg", "groomrate_stm", "groomrate_cst"
                  ))
  
  simulation <- parLapply(cl, 1:nrow(task_grid), run_one_parameter_set)

  on.exit(stopCluster(cl), add = TRUE)
})


saveRDS(simulation, "data/sim_100gs_10sims.Rdata")
```
Now, I need a way to sample from these data.

```{r calculate traits - total pop}
# read in simulation data 
# sim10 <- readRDS("data/simulation_10rep.Rdata")
# sim <- readRDS("data/simulation.Rdata")
# sim100 <- readRDS("data/simulation100.Rdata")


# optional flatten to dataframe
# results_sim10 <-  do.call(rbind, sim10)
# results_sim10$id <- paste(results_sim10$param_id, "-", results_sim10$sim_id, sep = "")

# create functions to calculate output values 

# FLEA POPULATION TRAITS - calculate relevant traits 
## Snapshot/Sampling Event: the count of fleas at one snapshot (e.g., end of day)

# function to calculate daily snapshot 

end_of_day_flea_pop <- function(sim){
  
  init_row <- data.frame(  # pre load initial conditions simulations 
    event = 0,
    time = 0,
    population = 1,
    event_type = "initial",
    sim_id = sim$sim_id[1],
    param_id = sim$param_id[1]
  )

  sim %>% # Apply operations to the dataset
    bind_rows(init_row, .) %>%
    mutate(time_unit = floor(time)) %>% # Convert time to integer units
    group_by(time_unit) %>%
    slice_tail(n = 1) %>% # Get the last observation per unit
    ungroup() %>%
    select(time_unit, population) %>% # Keep relevant columns
    
    # create full sequence
    tidyr::complete(time_unit = 0:780) %>%
    
    # carry previous value forward
    tidyr::fill(population, .direction = "down") %>% 
    dplyr::filter(time_unit != 0) %>% # get rid of the initial seeding
    pull(population) # only get a vector of the population sizes 
}

## values for one simulation

summarize_simulations <- function(simulation_list) {
  
  result <- lapply(seq_along(simulation_list), function(i) {
    
    eod_burdens <- end_of_day_flea_pop(simulation_list[[i]])
    
     eod_burden_list <-  list( # save list of the eod values for each sim 
       sim_id = i,
       eod_burdens = eod_burdens)


    sim_df <- simulation_list[[i]]
    
    
    data.frame(
      list_id = i,
      sim_id = unique(sim_df$sim_id)[1],
      param_id = unique(sim_df$param_id)[1],
      ind_metric_total_pop = sum(sim_df$event_type == "immigration") + # calculate total pop.size
                        sum(sim_df$event_type == "birth"),
      ind_metric_total_groom = sum(sim_df$event_type == "groomed"), # calculate total flea groomed
      
      ind_metric_infest_days = sum(eod_burdens  > 0),
      ind_metric_mean_daily_pop = mean(eod_burdens)
    )
  })
  
  do.call(rbind, result)
}

# run for each simulation 
system.time({
  flea_stat_df <- summarize_simulations(simulation) # each simulation
}) 

# summarize by parameters 
flea_stat_per_param <- flea_stat_df %>% 
  group_by(param_id) %>% 
  summarise(ind_metric_total_pop = mean(ind_metric_total_pop), 
            ind_metric_total_grm = mean(ind_metric_total_groom), 
            ind_metric_inf_days = mean(ind_metric_infest_days), 
            ind_metric_avg_dy_burden = mean(ind_metric_mean_daily_pop))

# merge in parameters 
param_grid$param_id <-  seq_along(tig_axes[,1])
flea_stat_per_param <- left_join(flea_stat_per_param, param_grid, by ="param_id")

# Save output 
write.csv(flea_stat_per_param, file ="data/exp1_output.csv")

# lets look at no grooming for now: 
## filter to the no grooming
flea_stat_per_param_ng <- flea_stat_per_param %>% 
  dplyr::filter(groom_strategy=="none")
## plot heatmap of total flea pop
ggplot(flea_stat_per_param_ng, 
       aes(x = theta, y = iota, fill = ind_metric_total_pop)) +
  geom_tile() +
  scale_y_continuous(breaks = seq(0, 0.5, by = 0.1)) +
  scale_fill_gradientn(colours = c("grey80", viridis::viridis(255, direction = -1)),
                       values = scales::rescale(c(1, 500)),
                       name = "total flea population", 
                       breaks = c(1,100,200,300,400,500),
                       labels = c("1","100","200","300","400","500")) +
  labs(x = "theta",
       y = "iota",
       fill = "pop") +
  theme_minimal()

# now for a mid grooming scenarios (50) plot, faceted by different strategies 
flea_stat_per_param_g51 <- flea_stat_per_param %>% 
  dplyr::filter(scenario==51)
## define levels of groom_stragety 
flea_stat_per_param_g51$groom_strategy <- factor(flea_stat_per_param_g51$groom_strategy, 
                                                 levels = c("none", "programmed", "stimulus",
                                                            "costly"))
               
## plot heatmap of total flea pop w variable t and i, but intermediate groom

intgroom_ti_heat <- ggplot(flea_stat_per_param_g51, 
       aes(x = theta, y = iota, fill = ind_metric_total_pop)) +
  geom_tile() +
  facet_wrap(~groom_strategy, ncol = 1) +
  scale_y_continuous(breaks = seq(0, 0.5, by = 0.1)) +
  scale_fill_gradientn(colours = c("grey80", viridis::viridis(255, direction = -1)),
                       values = scales::rescale(c(1, 1000)),
                       name = "total flea population", 
                       breaks = c(1,250,500,750,1000),
                       labels = c("1","250","500","750","1000")) +
  labs(x = "theta",
       y = "iota",
       fill = "pop") +
  theme_minimal()

png(filename="figures/FigureX3_midgroom-heat.png", width = 16, height =8, unit = 'in', res = 300)
intgroom_ti_heat
dev.off()

# plot individual infestation traits by resource 
## convert into long data 
flea_stat_per_param_long <- flea_stat_per_param %>%
  pivot_longer(cols = starts_with("ind_metric_"), 
                names_to = "metric",
                values_to = "value") 

# confirm groom strategy factors 
flea_stat_per_param_long$groom_strategy <- factor(flea_stat_per_param_long$groom_strategy,
                                                  levels = c("none", "programmed", "stimulus", "costly"))

# plot individual metrics for only resource-induced changes to grooming, and no immigration 
ggplot(flea_stat_per_param_long[flea_stat_per_param_long$iota == 0, ], 
       aes(x = scenario, y = value, 
           group = interaction(groom_strategy, theta), 
           color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(groom_strategy~metric, ncol = 4, nrow=4, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()

# plot individual metrics for only resource-induced changes to grooming, and baseline immigration 
rgroom_bi_plot <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$iota == 0.2, ], 
       aes(x = scenario, y = value, 
           group = interaction(groom_strategy, theta), 
           color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(groom_strategy~metric, ncol = 4, nrow=4, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()

# save r-grooming w baseline immigration
png(filename="figures/FigureX53_ind_metrics_rg-link_i02.png", width = 16, height =8, unit = 'in', res = 300)
rgroom_bi_plot
dev.off()


## Plot separate panels for individual metrics for only resource-induced changes to grooming and no immigration
totpop_plot_ni <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$iota == 0 & 
                                                  flea_stat_per_param_long$metric == "ind_metric_total_pop",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta),  
                          color = groom_strategy)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4,1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
totpop_plot_ni

# remove legend for multi-panel plot: 
totpop_plot_ni_nl <- totpop_plot_ni +
  theme(legend.position = "none",
        axis.title.x = element_blank()) # use joint label instead

## average eod burden 
dayburden_plot_ni <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$iota == 0 &
                                                               flea_stat_per_param_long$metric == 
                                                            "ind_metric_avg_dy_burden",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "mean daily burden") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
dayburden_plot_ni

# remove legend for multi-panel plot: 
dayburden_plot_ni_nl <- dayburden_plot_ni +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) # use joint label instead

## total fleas groomed  
grooming_effort_plot_ni <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$iota == 0 &
                                                             flea_stat_per_param_long$metric == 
                                                            "ind_metric_total_grm",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta),
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4,1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total groomed population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
grooming_effort_plot_ni

# remove legend for multi-panel plot: 
grooming_effort_plot_ni_nl <- grooming_effort_plot_ni +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none")
  
## infestation days 
inf_days_plot_ni <- ggplot(flea_stat_per_param_long[flea_stat_per_param_long$iota == 0 &
                                                      flea_stat_per_param_long$metric == 
                                                            "ind_metric_inf_days",], 
                      aes(x = scenario, y = round(value/780*100,2), 
                          group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "days with an infestation (%)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank()) # use joint label instead
inf_days_plot_ni # check all good 

# Multipanel plot for no immigration 
png(filename="figures/FigureX4_ind_metrics_ni.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(totpop_plot_ni_nl, dayburden_plot_ni_nl , grooming_effort_plot_ni_nl, inf_days_plot_ni, 
             nrow = 1,
             bottom = "resource availability")
dev.off()

# Now Resources jointly impacts immigration and grooming 


## convert so resources affects immigration same as grooming 
# calculate appropriate 
resource_pairs <- data.frame(
  scenario = c(1, 10,20,30,40,50,60,70,80,90,100),
  iota = c(0, 0.05, 0.1, 0.15,0.2,0.25,0.3,0.35,0.4,0.45,0.5)
)

flea_stat_per_param_long_ri_link <- flea_stat_per_param_long %>%
  inner_join(resource_pairs, by = c("scenario", "iota"))

# confirm grooming strategy levels
flea_stat_per_param_long_ri_link$groom_strategy <- factor(flea_stat_per_param_long_ri_link$groom_strategy,
                                                          levels = c("none", 
                                                                     "programmed", 
                                                                     "stimulus", 
                                                                     "costly")) 
# prepare individual metric plots 

# PLOT OF RESOURCE on individual metrics with immigration-grooming paired and Theta = 2
## total population 
totpop_plot <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$theta==2 &
                      flea_stat_per_param_long_ri_link$metric == "ind_metric_total_pop",], 
                      aes(x = scenario, group = groom_strategy, y = value, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
totpop_plot

# remove legend for multi-panel plot: 
totpop_plot_nl <- totpop_plot +
  theme(legend.position = "none",
        axis.title.x = element_blank()) # use joint label instead

## average eod burden 
dayburden_plot <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$theta==2 &
                                                            flea_stat_per_param_long_ri_link$metric == 
                                                            "ind_metric_avg_dy_burden",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "mean daily burden") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
dayburden_plot

# remove legend for multi-panel plot: 
dayburden_plot_nl <- dayburden_plot +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) # use joint label instead

## total fleas groomed  
grooming_effort_plot <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$theta==2 &
                                                            flea_stat_per_param_long_ri_link$metric == 
                                                            "ind_metric_total_grm",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total groomed population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
grooming_effort_plot

# remove legend for multi-panel plot: 
grooming_effort_plot_nl <- grooming_effort_plot +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none")
  
## infestation days 
inf_days_plot <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$theta==2 &
                                                            flea_stat_per_param_long_ri_link$metric == 
                                                            "ind_metric_inf_days",], 
                      aes(x = scenario, y = round(value/780*100,2), 
                          group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "days with an infestation (%)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank()) # use joint label instead
inf_days_plot # check all good 

# Multipanel plot for one theta value (2)
png(filename="figures/FigureX52_ind_metrics_ri-link_t2.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(totpop_plot_nl, dayburden_plot_nl , grooming_effort_plot_nl, inf_days_plot, nrow = 1,
             bottom = "resource availability")
dev.off()

# PLOT OF RESOURCE on individual metrics with immigration-grooming paired with variable Theta
## total population for all thetas
totpop_plot_at <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$metric == 
                                                            "ind_metric_total_pop",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_x_continuous(limits = c(1,100)) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
totpop_plot_at 

# remove legend for multi-panel plot: 
totpop_plot_at_nl <- totpop_plot_at +
  theme(legend.position = "none",
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## average eod burden for all thetas
dayburden_plot_at <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$metric == 
                                                            "ind_metric_avg_dy_burden",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "mean daily burden") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
dayburden_plot_at

# remove legend for multi-panel plot: 
dayburden_plot_at_nl <- dayburden_plot_at +
  theme(legend.position = "none", 
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## total fleas groomed  
grooming_effort_plot_at <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$metric == 
                                                            "ind_metric_total_grm",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy,theta),
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total groomed population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
grooming_effort_plot_at

# remove legend for multi-panel plot: 
grooming_effort_plot_at_nl <- grooming_effort_plot_at +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none",
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
  
## infestation days 
inf_days_plot_at <- ggplot(flea_stat_per_param_long_ri_link[flea_stat_per_param_long_ri_link$metric == 
                                                            "ind_metric_inf_days",], 
                      aes(x = scenario, y = round(value/780*100,2), 
                          group = interaction(groom_strategy,theta),  
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "days with an infestation (%)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank(),# use joint label instead
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
inf_days_plot_at # check all good 

# Save Plot 
png(filename="figures/FigureX5_ind_metrics_ri-link.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(totpop_plot_at_nl, dayburden_plot_at_nl , grooming_effort_plot_at_nl, 
                                  inf_days_plot_at, nrow = 1,
             bottom = grid::textGrob("resource availability",
                               gp = grid::gpar(fontsize = 20, fontface = "bold")))
dev.off()
```

```{r subsetting the eod }

# create a function to sample only a portion of the days (e.g. last 25%)
subset_simulation <- function(sim_df, range_percent) {
  # Ensure range_percent is a valid numeric vector of length 2 (e.g., c(0, 10) or c(75, 100))
  if (length(range_percent) != 2 || any(range_percent < 0) || any(range_percent > 100)) {
    stop("range_percent must be a numeric vector of length 2 between 0 and 100.")
  }
  
  # Get the start and end time from the simulation
  min_time <- min(sim_df$time_unit)
  max_time <- max(sim_df$time_unit)

  # Compute the time range corresponding to the percentage range
  time_start <- min_time + (range_percent[1] / 100) * (max_time - min_time)
  time_end <- min_time + (range_percent[2] / 100) * (max_time - min_time)

  # Subset the dataframe based on the calculated time range
  sim_df %>%
    filter(time_unit >= time_start & time_unit <= time_end)
}

# limit population snapshots to latter 75% (eliminating burn-in period)
# sim_eod_pop_last75 <- lapply(eod_burden_list, subset_simulation, range_percent = c(26, 100))

# Create master dataframe (likely not needed)
# sim_eod_pop_last75_combined <- bind_rows(
#  lapply(seq_along(sim_eod_pop_last75), function(i) {
#    sim_eod_pop_last75[[i]] %>% mutate(sim_id = as.factor(i))  # Assign simulation ID
#  })
# )
```

```{r population sim}
# Population experiment 
# overwrite for population
num_simulations <- 1  # Number of times to run the simulation - this time just once very long
max.time <- 3650 # 10 years -- > beyond on squirrel lifetime 
# Define the task grid: 
task_grid <- expand.grid(
    param_id = 1:nrow(param_grid), # using same parameters
    sim_id   = 1:num_simulations)

# run the simulation for the population 
system.time({

  cl <- makeCluster(10)
  
  # Move libraries to the clusters 
  clusterEvalQ(cl, library(tidyverse))

  # Move main environment items to each of the clusters
  clusterExport(cl, 
                c("groom_prg_scenarios","groom_stm_scenarios", "groom_cst_scenarios",
                   "param_grid","max.time", "task_grid", 
                   "birthrate", "deathrate_dd", "immigrationrate", 
                  "groom_strategies", "get_groom_rate", "groomrate_prg", "groomrate_stm", "groomrate_cst"
                  ))
  
  simulation_pop <- parLapply(cl, 1:nrow(task_grid), run_one_parameter_set)

  on.exit(stopCluster(cl), add = TRUE)
})

# save output 
saveRDS(simulation_pop, "data/simulation_pop.Rdata")
```

```{r calculate host population traits}
# calculate the autocorrelation and the needed space between pops 
# find_acf_gap <- function(x, cutoff = 0.1, lag.max = 200) {
#  a <- acf(x, plot = FALSE, lag.max = lag.max, na.action = na.pass)
#  lags <- a$lag[-1]
#  acfs <- a$acf[-1]
  
#  idx <- which(abs(acfs) < cutoff)[1]
  
#  if (is.na(idx)) return(NA)
#  lags[idx]
# }

# find_acf_gap(eod_burdens_list[[1]]$population)

# Calculate simulation values 
# create a list of time points to get the values 
sample_days <- seq(949, 3449, by = 25) # last 75% of 10 years, every 25 days apart 


# update eod function for full simulation: 
end_of_day_flea_pop_3650 <- function(sim){
  
  init_row <- data.frame(  # pre load initial conditions simulations 
    event = 0,
    time = 0,
    population = 1,
    event_type = "initial",
    sim_id = sim$sim_id[1],
    param_id = sim$param_id[1]
  )

  sim %>% # Apply operations to the dataset
    bind_rows(init_row, .) %>%
    mutate(time_unit = floor(time)) %>% # Convert time to integer units
    group_by(time_unit) %>%
    slice_tail(n = 1) %>% # Get the last observation per unit
    ungroup() %>%
    select(time_unit, population) %>% # Keep relevant columns
    
    # create full sequence
    tidyr::complete(time_unit = 1:3650) %>%
    
    # carry previous value forward
    tidyr::fill(population, .direction = "down") %>% 
    dplyr::filter(time_unit >= 1, time_unit <= 3650) %>% # get rid of the initial seeding
    pull(population) # only get a vector of the population sizes 
}


## function to sample across the simulation
sample_snapshots_one_pop <- function(simulation_list, sample_days, i) {
  
    pop_eod_burdens <- end_of_day_flea_pop_3650(simulation_list[[i]])
    
    pop_eod_burden_list <-  list( # save list of the eod values for each sim 
       sim_id = i,
       eod_burdens = pop_eod_burdens)
    
    pop_eod_burdens_df <- data.frame(time_unit = c(1:3650), eod_burden = pop_eod_burdens)
    
    pop_eod_burden_last75 <- subset_simulation(pop_eod_burdens_df,
                                               range_percent = c(26, 100))
    pop_sampled_burden <- pop_eod_burden_last75[pop_eod_burden_last75$time_unit %in% sample_days, ]
    pop_sampled_burden
}

# load sim_pop if needed 
# simulation_pop <- load("data/simulation_pop.Rdata")
# create a list of the sampled day values per pop sim 
pop_sampled_burden_list <- lapply(
  seq_along(simulation_pop),
  function(i) sample_snapshots_one_pop(simulation_pop, sample_days, i)
)

# Now summarize for these 100 samples 
summarize_pop_simulations <- function(simulation_list) {
  
  result <- lapply(seq_along(simulation_list), function(i) {
    
    # save output as a dataframe 
    
    pop_sim_df <- simulation_list[[i]]
    
    pop_metric_prev <- sum(pop_sim_df$eod_burden > 0)
    pop_metric_abun <- sum(pop_sim_df$eod_burden) / 100
    pop_metric_inte <- pop_metric_abun / pop_metric_prev
      
    mu <- mean(pop_sim_df$eod_burden)
    v <- var(pop_sim_df$eod_burden)
    k <- mu^2 / (v - mu)
      
    pop_metric_crowd <- mu + mu / k
    pop_metric_patch <- 1 + 1 / k
    
    data.frame(
      list_id = i,
      param_id = i,
      pop_metric_prev = pop_metric_prev,
      pop_metric_abun = pop_metric_abun,
      pop_metric_inte = pop_metric_inte,
      mu = mu,
      v = v,
      pop_metric_k = k,
      pop_metric_crowd = pop_metric_crowd,
      pop_metric_patch = pop_metric_patch
    )
  }) 
  
  do.call(rbind, result)
}

# run for each simulation 
system.time({
  pop_stat_df <- summarize_pop_simulations(pop_sampled_burden_list) # each simulation
}) 

```

```{r plot population measures across resource scenarios }
# merge in parameters 
pop_stat_per_param <- left_join(pop_stat_df, param_grid, by ="param_id")

# save the pop eod spaced burdens
saveRDS(pop_stat_per_param, "data/sim_pop_metrics.Rdata")

# correct abundance to mean abundance 
pop_stat_per_param$pop_metric_abun <- pop_stat_per_param$pop_metric_abun / 100

# lets look at no grooming for now: 
## filter to the no grooming
pop_stat_per_param_ng <- pop_stat_per_param %>% 
  dplyr::filter(groom_strategy=="none")
## plot heatmap of prevalence
prev_it_ng_plot <- ggplot(pop_stat_per_param_ng, 
       aes(x = theta, y = iota, fill = pop_metric_prev)) +
  geom_tile() +
  scale_y_continuous(breaks = seq(0, 0.5, by = 0.1)) +
  scale_fill_gradientn(colours = c("grey80", viridis::viridis(255, direction = -1)),
                       values = scales::rescale(c(1, 101)),
                       name = "population prevalence", 
                       breaks = c(0,25,50,75,100),
                       labels = c("1","25","50","75","100")) +
  labs(x = "theta",
       y = "iota",
       fill = "prevalence") +
  theme_minimal()
prev_it_ng_plot # check all good

png(filename="figures/FigureX_ti_ng_prev.png", width = 16, height =8, unit = 'in', res = 300)
prev_it_ng_plot
dev.off()

# now for a mid grooming scenarios (50) plot, faceted by different strategies 
pop_stat_per_param_g51 <- pop_stat_per_param %>% 
  dplyr::filter(scenario==51)
## define levels of groom_stragety 
pop_stat_per_param_g51$groom_strategy <- factor(pop_stat_per_param_g51$groom_strategy, 
                                                 levels = c("none", "programmed", "stimulus",
                                                            "costly"))
               
## plot heatmap of prevalence
ti_midg_plot <- ggplot(pop_stat_per_param_g51, 
       aes(x = theta, y = iota, fill = pop_metric_prev)) +
  geom_tile() +
  facet_wrap(~groom_strategy, ncol = 1) +
  scale_y_continuous(breaks = seq(0, 0.5, by = 0.1)) +
  scale_fill_gradientn(colours = c("grey80", viridis::viridis(255, direction = -1)),
                       values = scales::rescale(c(1, 101)),
                       name = "prevalence", 
                       breaks = c(0,25,500,75,100),
                       labels = c("0","25","50","75","100")) +
  labs(x = "theta",
       y = "iota",
       fill = "prevalence") +
  theme_minimal()

ti_midg_plot # check all good 

png(filename="figures/FigureX_ti_midg_prev.png", width = 16, height =8, unit = 'in', res = 300)
ti_midg_plot
dev.off()

# plot individual infestation traits by resource 
## convert into long data 
pop_stat_per_param_long <- pop_stat_per_param %>%
  pivot_longer(cols = starts_with("pop_metric_"), 
                names_to = "metric",
                values_to = "value") 

# confirm groom strategy factors 
pop_stat_per_param_long$groom_strategy <- factor(pop_stat_per_param_long$groom_strategy,
                                                  levels = c("none", 
                                                             "programmed", 
                                                             "stimulus",
                                                             "costly"))

# plot individual metrics for only resource-induced changes to grooming, and no immigration 

pop_met_rg_ni_plot <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0, ], 
       aes(x = scenario, y = value, 
           group = interaction(groom_strategy, theta), 
           color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(groom_strategy~metric, ncol = 4, nrow=6, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()
pop_met_rg_ni_plot# check all good 

png(filename="figures/FigureX_rg_ni_prev.png", width = 16, height =8, unit = 'in', res = 300)
pop_met_rg_ni_plot 
dev.off()

# plot individual metrics for only resource-induced changes to grooming, and baseline immigration (0.2)

pop_met_rg_bi_plot <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0.2, ], 
       aes(x = scenario, y = value, 
           group = interaction(groom_strategy, theta), 
           color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(metric ~groom_strategy, ncol = 4, nrow=6, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()

pop_met_rg_bi_plot # check all good

# # plot individual metrics for only resource-induced changes to grooming, and baseline immigration (0.2) and mid theta (2) 

pop_met_rg_bimt_plot <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0.2 &
                                                       pop_stat_per_param_long$theta == 5, ], 
       aes(x = scenario, y = value, 
           color = groom_strategy)) +
  geom_point(size = 2) +
  facet_wrap(metric~groom_strategy, ncol = 4, nrow=6, scales = "free_y") +
  labs(x = "resource availability",
       y = "value") +
  theme_minimal()

pop_met_rg_bimt_plot # check all good 

# save plot 
png(filename="figures/FigureX_rg_bimt_pop_metrics.png", width = 16, height =8, unit = 'in', res = 300)
pop_met_rg_bimt_plot 
dev.off()


## Plot separate panels for individual metrics for only resource-induced changes to grooming and no immigration
prev_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 & 
                                               pop_stat_per_param_long$metric == "pop_metric_prev",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta),  
                          color = groom_strategy)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4,1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "prevalence") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
prev_plot_ni

# remove legend for multi-panel plot: 
prev_plot_ni_nl <- prev_plot_ni +
  theme(legend.position = "none",
        axis.title.x = element_blank()) # use joint label instead

## average eod burden 
abund_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 &
                                                  pop_stat_per_param_long$metric ==
                                                  "pop_metric_avg_abund",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "abundance") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
abund_plot_ni

# remove legend for multi-panel plot: 
abund_plot_ni_nl <- abund_plot_ni +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) # use joint label instead

## Intensity  
intense_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 & 
                                                    pop_stat_per_param_long$metric == 
                                                            "pop_metric_inte",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta),
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4,1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "total groomed population") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
intense_plot_ni

# remove legend for multi-panel plot: 
intense_plot_ni_nl <- intense_plot_ni +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none")
  
## k aggregation
k_plot_ni <- ggplot(pop_stat_per_param_long[pop_stat_per_param_long$iota == 0 &
                                            pop_stat_per_param_long$metric == "pop_metric_k",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 1) +
  labs(x = "resource availability",
       y = "aggegation (k)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank()) # use joint label instead
k_plot_ni # check all good 

# Multipanel plot for no immigration 
png(filename="figures/FigureX6_pop_metrics_ni.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(prev_ni_nl, abund_plot_ni_nl , intense_plot_ni_nl, k_plot_ni, 
             nrow = 1,
             bottom = "resource availability")
dev.off()

# Now Resources jointly impacts immigration and grooming  
## use same resource_pairs as individual experiment

pop_stat_per_param_long_ri_link <- pop_stat_per_param_long %>%
  inner_join(resource_pairs, by = c("scenario", "iota"))

# confirm grooming strategy levels
pop_stat_per_param_long_ri_link$groom_strategy <- factor(pop_stat_per_param_long_ri_link$groom_strategy,
                                                          levels = c("none", 
                                                                     "programmed", 
                                                                     "stimulus", 
                                                                     "costly")) 
# prepare individual metric plots 

# PLOT OF RESOURCE on individual metrics with immigration-grooming paired and Theta = 5
## total population 
prev_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                        pop_stat_per_param_long_ri_link$metric == "pop_metric_prev",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "prevalence") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
prev_plot

# remove legend for multi-panel plot: 
prev_plot_nl <- prev_plot +
  theme(legend.position = "none",
        axis.title.x = element_blank()) # use joint label instead

## average eod burden 
abund_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                                    pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_abun",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean abundance") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
abund_plot

# remove legend for multi-panel plot: 
abund_plot_nl <- abund_plot +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) # use joint label instead

## INTENSITY  
intense_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                                       pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_inte",], 
                      aes(x = scenario, y = value, group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean intensity") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
intense_plot

# remove legend for multi-panel plot: 
intense_plot_nl <- intense_plot +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none")
  
## Aggregation k 
k_plot <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$theta==7 &
                                                 pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_k",], 
                      aes(x = scenario, y = round(value/780*100,2), 
                          group = groom_strategy, color = groom_strategy)) +
  geom_line(size = 2) +
  
  facet_wrap(~groom_strategy, ncol = 4) +
  geom_line(y= 1, color = "black") +
  geom_line(y= 0, color = "grey") +
  geom_line(y= 20, color = "grey") +
  scale_y_continuous(limits = c(0,20)) +
  labs(x = "resource availability",
       y = "aggregation (k)") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank()) # use joint label instead
k_plot # check all good 

# Multipanel plot for one theta value (5)
png(filename="figures/FigureX_pop_metrics_rgi_t7.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(prev_plot_nl, abund_plot_nl , intense_plot_nl, k_plot, nrow = 4,
             bottom = "resource availability")
dev.off()

# PLOT OF RESOURCE on individual metrics with immigration-grooming paired with variable Theta
## total population for all thetas
prev_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_prev",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "prevalence") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
prev_plot_at 

# remove legend for multi-panel plot: 
prev_plot_at_nl <- prev_plot_at +
  theme(legend.position = "none",
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## average eod burden for all thetas
abund_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_abun",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy, theta), 
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean abundance") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
abund_plot_at

# remove legend for multi-panel plot: 
abund_plot_at_nl <- abund_plot_at +
  theme(legend.position = "none", 
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14),
        axis.title.x = element_blank()) # use joint label instead

## Intensity  
intense_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_inte",], 
                      aes(x = scenario, y = value, group = interaction(groom_strategy,theta),
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "mean intensity") +
  theme_minimal()  +
  theme(strip.text.x = element_blank())
intense_plot_at

# remove legend for multi-panel plot: 
intense_plot_at_nl <- intense_plot_at +
  theme(axis.title.x = element_blank(), # use joint label instead 
        legend.position = "none",
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
  
## k aggregation 
k_plot_at <- ggplot(pop_stat_per_param_long_ri_link[pop_stat_per_param_long_ri_link$metric == 
                                                            "pop_metric_k",], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy,theta),  
                          color = groom_strategy, alpha = theta)) +
  geom_line(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  scale_y_continuous(limits = c(0,5)) +
  facet_wrap(~groom_strategy, ncol = 4) +
  labs(x = "resource availability",
       y = "aggregation (k)") +
  geom_line(y= 1, color = "black") +
  geom_line(y= 0, color = "grey") +
  geom_line(y= 20, color = "grey") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank(),# use joint label instead
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
k_plot_at  # check all good 

# Save Plot 
png(filename="figures/FigureX7_pop_metrics_ri-link.png", width = 16, height =8, unit = 'in', res = 300)
grid.arrange(prev_plot_at_nl, abund_plot_at_nl, intense_plot_at_nl, k_plot_at, nrow = 4,
             bottom = grid::textGrob("resource availability",
                               gp = grid::gpar(fontsize = 20, fontface = "bold")))
dev.off()

# Lets focus on just costly and stimulus, as they appear to only have suitable k variation, limit theta 
k_plot_at_sc_3t <- ggplot(pop_stat_per_param_long_ri_link[
  pop_stat_per_param_long_ri_link$metric == "pop_metric_k" &
  pop_stat_per_param_long_ri_link$groom_strategy %in% c("stimulus", "programmed") &
  pop_stat_per_param_long_ri_link$theta == 7,], 
                      aes(x = scenario, y = value, 
                          group = interaction(groom_strategy,theta),  
                          color = groom_strategy, alpha = theta)) +
  geom_point(size = 2) +
  scale_alpha(range = c(0.4, 1)) +
  scale_y_continuous(limits = c(0,5)) +
  facet_wrap(~groom_strategy, ncol = 2) +
  labs(x = "resource availability",
       y = "aggregation (k)") +
  geom_line(y= 1, color = "black") +
  geom_line(y= 0, color = "grey") +
  geom_line(y= 20, color = "grey") +
  theme_minimal()  +
  theme(strip.text.x = element_blank(), 
        axis.title.x = element_blank(),# use joint label instead
        axis.title.y = element_text(face = "bold", size =20),
        axis.text.x = element_text(size=14),
        axis.text.y= element_text(size=14))
k_plot_at_sc_3t 
```